---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
require(pipeflow)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  prompt = F,
  fig.path = "images/README-",
  tidy = FALSE,
  cache = FALSE,
  out.width = "100%"
)

old <- options(width = 100L)
```

<!-- badges: start -->

[![test-coverage](https://github.com/rpahl/pipeflow/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/rpahl/pipeflow/actions/workflows/test-coverage.yaml)
[![R-CMD-check](https://github.com/rpahl/pipeflow/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/rpahl/pipeflow/actions/workflows/check-standard.yaml)
[![lint](https://github.com/rpahl/pipeflow/actions/workflows/lint.yaml/badge.svg)](https://github.com/rpahl/pipeflow/actions/workflows/lint.yaml)

<!-- badges: end -->

The goal of `pipeflow` is to provide a *lightweight* yet *powerful*
framework for composing and managing custom data analysis pipelines.
Dependencies among analysis steps are tracked automatically and are
used to re-run only steps that are affected by input or parameter
changes, similar to the `make` utility.

The package provides similar functionality to the
[targets](https://CRAN.R-project.org/package=targets) package, but
aims to be more *lightweight* and *easier* to use and understand.
It can be used in a wide range of
data analysis projects, from small exploratory analyses to large
production pipelines while still providing a high level of control
over the analysis process.

### When to use `pipeflow`

* first reason
* second reason
* third reason


### Installation

```{r, eval = FALSE}
# Install release version from CRAN
install.packages("pipeflow")

# Install development version from GitHub
devtools::install_github("rpahl/pipeflow")
```


### Usage

```{r}
library(pipeflow)
```

### Getting Started


```{r, include = FALSE}
options(old)
```