---
title: "Combining pipelines"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to combine different pipelines to a single pipeline.
vignette: >
  %\VignetteIndexEntry{Combining pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
require(pipeflow)

knitr::opts_chunk$set(
  comment = "#",
  prompt = FALSE,
  tidy = FALSE,
  cache = FALSE,
  collapse = TRUE
)

old <- options(width = 100L)
library(ggplot2)
```

The possibility to combine pipelines basically allows to modularize the pipeline
creation process. This is especially useful when you have a set of pipelines that are
used in different contexts and you want to avoid code duplication.

In this vignette we will also introduce the pipeflow alias functions, that is,
for each member function of a pipeline, there is an alias function, which allows to
create chain pipeline steps using R's native pipe operator `|>`.
For example, the `add` function has an alias `pipe_add` (see below).


### Define two pipelines

Let's define one pipeline that is used for data_preprocessing and one that does the
modeling.

Data preprocessing pipeline:
```{r define-prepocessing-pipeline}
library(pipeflow)
library(ggplot2)

pip1 <- pipe_new(
        "preprocessing",
        data = airquality
    ) |>

    pipe_add(
        "data_prep",
        function(data = ~data) {
            replace(data, "Temp.Celsius", (data[, "Temp"] - 32) * 5/9)
        }
    ) |>

    pipe_add(
        "standardize",
        function(
            data = ~data_prep,
            yVar = "Ozone"
        ) {
            data[, yVar] <- scale(data[, yVar])
            data
        }
    )
```

```{r}
pip1
```

Modeling pipeline:
```{r define-modeling-pipeline}
pip2 <- pipe_new(
        "modeling",
        data = airquality
    ) |>

    pipe_add(
        "fit",
        function(
            data = ~data,
            xVar = "Temp",
            yVar = "Ozone"
        ) {
            lm(paste(yVar, "~", xVar), data = data)
        }
    ) |>

    pipe_add(
        "plot",
        function(
            model = ~fit,
            data = ~data,
            xVar = "Temp",
            yVar = "Ozone",
            title = "Linear model fit"
        ) {
            coeffs <- coefficients(model)
            ggplot(data) +
                geom_point(aes(.data[[xVar]], .data[[yVar]])) +
                geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
                labs(title = title)
        }
    )
```


```{r}
pip2
```

Graphically, the two pipelines look as follows:

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip1$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip2$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

### Combine pipelines

Next we combine the two pipelines. We can do this by using the `append` function.

```{r}
pip <- pip1$append(pip2)

pip
```

First of all, note that the `data` step of the second pipeline
has been appended with the name of the second pipeline. In particular, the first
step of the second pipeline has been renamed from `data` to `data.modeling`
(line 4 in the `step` column) and likewise the dependencies of the second pipeline
have been updated (see lines 5-6 in the `depends` column).

That is, when appending two pipelines, `pipeflow` ensures that the
step names remain unique in the resulting combined pipeline and therefore
automatically renames duplicated step names if necessary.

Now, as can be also seen from the graphical representation of the pipeline,

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 250))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

the two pipelines are not yet connected. To make actual use of the combined
pipeline, we therefore want to use the output of the first pipeline as input of
the second pipeline, that is, we want to use the output of the
`standardize` step as the data parameter input in the `data.modeling` step.
One way to achieve this would be to use the `replace` function as described earlier
in the vignette [modify the pipeline](modify-pipeline.html), for example:

```{r}
pip$replace_step("data.modeling", function(data = ~standardize) data)

pip
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

#### Relative indexing

Since the name of the last step might not always be known^[A typical example would
be appending several pipelines in a programmatic context.], the `pipeflow` package
also provides a relative position indexing mechanism, which allows to rewrite the
above command as follows:

```{r}
pip$replace_step("data.modeling", function(data = ~-1) data)

pip
```

Generally speaking, the relative indexing mechanism allows to refer to steps positioned
above the current step. The index `~-1` can be interpreted as "go one step back", `~-2`
as "go two steps back", and so on.

Since the scenario of connecting two pipelines is so common and to avoid having to do
the above replacement steps manually, the `append` function actually provides an
argument `outAsIn` that allows for appending and "connecting" both pipelines in one go:

```{r}
pip <- pip1$append(pip2, outAsIn = TRUE)

pip
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

If we inspect the `data.modeling` step, we see that "under the hood" the original
step indeed has been replaced by the output of the last step of the first pipeline using
the same relative indexing mechanism we did manually before.

```{r}
pip$get_step("data.modeling")[["fun"]]
```

### Run combined pipeline

Let's now run the combined pipeline and inspect the plot.

```{r}
pip$run()
```

```{r, fig.alt = "model-plot"}
pip$get_out("plot")
```

As we can see, the plot shows the linear model fit using the standardized data.
We can now go ahead and for example change the x-variable of the model and rerun the
pipeline.

```{r}
pip$set_params(list(xVar = "Temp.Celsius"))
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

```{r}
pip$run()
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

```{r, fig.alt = "model-plot"}
pip$get_out("plot")
```

When creating these pipelines, usually there will be a lot of steps calculating
intermediate results and only a few steps contain the final results that we are
interested in. In the above example, we were interested in the final plot
output.
In a real-world scenario, the pipeline would contain many more steps that are not
of interest to us.
To see how to conveniently tag, collect and possibly group the output of those
final steps, see the next vignette [Collecting output](collect-output.html).

```{r, include = FALSE}
options(old)
```
