---
title: "Get started with pipeflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Start here if this is your first time using pipeflow.
vignette: >
  %\VignetteIndexEntry{Get started with pipeflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
require(pipeflow)

knitr::opts_chunk$set(
  comment = "#",
  prompt = FALSE,
  tidy = FALSE,
  cache = FALSE,
  collapse = TRUE
)

old <- options(width = 100L)
```

## A simple example to get started
In this example, we'll use base R's airquality dataset.

```{r show-airquality}
head(airquality)
```

Our goal is to create an analysis pipeline that performs the following steps:

* add new data column `Temp.Celsius` containing the temperature in degrees
  Celsius
* fit a linear model to the data
* plot the data and the model fit.

In the following, we'll show how to define and run the pipeline, how to inspect
the output of specifics steps, and finally how to re-run the pipeline with
different parameter settings, which is one of the selling points of using
such a pipeline.

### Define the pipeline

For easier understanding, we go step by step. First, we create a new pipeline
with the name "my-pipeline" and the airquality dataset as input data.

```{r define-pipeline}
library(pipeflow)
library(ggplot2)

pip <- Pipeline$new("my-pipeline", data = airquality)
```

Let's view the pipeline in it's table form after initialization.
```{r show-initial-pipeline}
pip
```

Here, each step is represented by one row in the table where the
`step` column lists the name of the step and the `state` column shows the
current state of the step. For now, don't worry about the other columns -
they are explained later.

As you can see, the data is always the first step in the pipeline and
initially it is in state `New`, which means the pipeline has not been run
yet.

Next, we add a step called `data_prep`, which consists of a function that
takes the output of the `data` step as its first argument, adds a new column
and returns the modified data as its output. To refer to the output of an
earlier pipeline step, we always use the tilde (~) operator.

```{r define-data-prep-step}
pip$add(
    "data_prep",
    function(data = ~data) {
        replace(data, "Temp.Celsius", (data[, "Temp"] - 32) * 5/9)
    }
)
```


Again, let's inspect the pipeline.
```{r show-pipeline-after-data-prep-was-added}
pip
```

A second step called `data_prep` was added and it depends on the `data`
step as can be seen in column `depends`, which will list all dependencies of
a step.

Next, we add a step called `model_fit`, which fits a linear model to the
data. The function takes the output of the `data_prep` and defines a
parameter `xVar`, which is used to specify the variable that is used as
predictor in the linear model.

```{r}
pip$add(
    "model_fit",
    function(
        data = ~data_prep,
        xVar = "Temp.Celsius"
    ) {
        lm(paste("Ozone ~", xVar), data = data)
    }
)
```

Lastly, we add a step called `model_plot`, which plots the data and the
linear model fit. The function uses the output from both the
`model_fit` and the `data_prep` step. It also defines the `xVar`
parameter and a parameter `title`, which is used as the title
of the plot.

```{r}
pip$add(
    "model_plot",
    function(
        model = ~model_fit,
        data = ~data_prep,
        xVar = "Temp.Celsius",
        title = "Linear model fit"
    ) {
        coeffs <- coefficients(model)
        ggplot(data) +
            geom_point(aes(.data[[xVar]], .data[["Ozone"]])) +
            geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
            labs(title = title)
    }
)
```

This completes our pipeline, which now looks as follows:
```{r show-completed-pipeline}
pip
```

In the last line, we see that the `model_plot` step depends on both
the `model_fit` and `data_prep` step.
To visualize a pipeline, we can use the `get_graph()` method, which returns
a list of arguments that can be passed to the `visNetwork` function from the
`visNetwork` package.

```{r, eval = FALSE}
library(visNetwork)
do.call(visNetwork, args = pip$get_graph()) |>
    visHierarchicalLayout(direction = "LR")
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR")
```

### Run the pipeline and inspect the output

To run the pipeline, we simply call the `run()` method on the pipeline,
which will produce the following output:

```{r run-pipeline}
pip$run()
```

Let's inspect the pipeline again.

```{r pipeline-after-run}
pip
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR")
```

We can see that the `state` of all steps have been changed from `New` to `Done`,
which graphically is represented by the color change from blue to green.

In addition, the output was added in the `out` column. We can inspect the
output of a specific step by calling the `get_out()` method on the pipeline.

```{r inspect-lm, message = FALSE}
pip$get_out("model_fit")
```

```{r inspect-plot, message = FALSE, warning = FALSE}
pip$get_out("model_plot")
```


### Change parameters of the pipeline

Any parameter of the pipeline that is *not* bound to a specific step
(i.e. is not defined with the tilde (~) operator) can be
changed by calling the `set_params()` method on the pipeline. Let's first
list all unbound parameters:

```{r inspect-params}
pip$get_params()
```

Now let's assume we want to change the `xVar` in the entire pipeline to
`"Solar.R"`. We can do this simply as follows:

```{r set-xVar}
pip$set_params(list(xVar = "Solar.R"))
pip$get_params()
```

Changing the parameters of the pipeline at least has an effect on all
steps that depend on the changed parameter.

```{r show-pipeline-with-outdated-step}
pip
```

```{r, echo = FALSE}
library(visNetwork)
do.call(visNetwork, args = c(pip$get_graph(), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR")
```

We can see that the `model_fit` and `model_plot` steps are now in state
`Outdated` (graphically indicated by the orange color), which means that they
need to be rerun. We can do this by
calling the `run()` method on the pipeline again.

```{r run-pipeline-again}
pip$run()
```


The outdated steps were re-run as expected and the output was
updated accordingly.

```{r inspect-plot-again, message = FALSE, warning = FALSE}
pip$get_out("model_plot")
```

To just change the title of the plot, obviously only the `model_plot` step
needs to be rerun.

```{r set-title}
pip$set_params(list(title = "Some new title"))
pip
```

```{r inspect-plot-after-title-change, message = FALSE, warning = FALSE}
pip$run()$get_out("model_plot")
```

If instead we were to change the input data, since all steps depend on it,
we expect all steps to be rerun.

```{r}
pip$set_data(airquality[1:10, ])
pip
```

```{r inspect-plot-after-data-change, message = FALSE, warning = FALSE}
pip$run()$get_out("model_plot")
```

Next, let's see how to [modify the pipeline](modify-pipeline.html).

```{r, include = FALSE}
options(old)
```
