---
title: "Modifying existing pipelines"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to insert, replace, and remove steps in a pipeline.
vignette: >
  %\VignetteIndexEntry{Modifying existing pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
library(pipeflow)

knitr::opts_chunk$set(
  comment = "#",
  prompt = FALSE,
  tidy = FALSE,
  cache = FALSE,
  collapse = TRUE
)

old <- options(width = 100L)
library(ggplot2)
```

### Existing pipeline

```{r define-pipeline, include = FALSE, echo = FALSE}
pip <- pip_new("my-pip")

pip |> pip_add("data", function(data = airquality[1:10, ]) data) |>

    pip_add("data_prep",
        function(x = ~data) {
            replace(x, "Temp.Celsius", (x[, "Temp"] - 32) * 5/9)
        }
    ) |>

    pip_add("model_fit",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius"
        ) {
            lm(paste("Ozone ~", xVar), data = data)
        }
    ) |>

    pip_add("model_plot",
        function(
            model = ~model_fit,
            data = ~data_prep,
            xVar = "Temp.Celsius",
            xLab = "Temperature in degrees Celsius",
            title = "Linear model fit"
        ) {
            require(ggplot2, quietly = TRUE)
            coeffs <- coefficients(model)
            ggplot(data) +
                geom_point(aes(.data[[xVar]], .data[["Ozone"]])) +
                geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
                labs(title = title, x = xLab)
        }
    )

pip_set_params(
    pip,
    list(
        xVar = "Solar.R",
        xLab = "Solar radiation in Langleys",
        title = "Some new title"
    )
)

pip_run(pip, lgr = NULL)
```

Let's start where we left off in the
[Get started with pipeflow](v01-get-started.html) vignette, that is, we have
the following pipeline

```{r show-pipeline}
pip
```

with the following set data

```{r show-data}
pip_get_params(pip)[["data"]] |> head(3)
```


### Insert new step

Let's say we want to insert a new step after the `data_prep` step
that standardizes the y-variable.

```{r insert-step}
pip |> pip_add("standardize",
    function(
        data = ~data_prep,
        yVar = "Ozone"
    ) {
        data[, yVar] <- scale(data[, yVar])
        data
    },
    after = "data_prep"
)
```


```{r}
pip
```

```{r, eval = FALSE, echo = nzchar(Sys.getenv("IN_PKGDOWN"))}
library(visNetwork)
do.call(visNetwork, args = pip_get_graph(pip)) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

```{r, echo = FALSE, eval = nzchar(Sys.getenv("IN_PKGDOWN"))}
library(visNetwork)
do.call(visNetwork, args = c(pip_get_graph(pip), list(height = 300))) |>
    visHierarchicalLayout(direction = "LR", sortMethod = "directed")
```

The `standardize` step is now part of the pipeline, but
so far it is not used by any other step.

### Replace existing steps

Let's revisit the function definition of the `model_fit` step

```{r}
pip[["model_fit", "fun"]]
```

To use the standardized data, we need to change the data dependency
such that it refers to the `standardize` step. Also instead of
a fixed y-variable in the model, let's pass it as a parameter.

```{r replace-model-fit-step}
pip |> pip_replace("model_fit",
    function(
        data = ~standardize,        # <- changed data reference
        xVar = "Temp.Celsius",
        yVar = "Ozone"              # <- new y-variable
    ) {
        lm(paste(yVar, "~", xVar), data = data)
    }
)
```


The `model_plot` step needs to be updated in a similar way.

```{r replace-model-plot-step}
pip |> pip_replace("model_plot",
    function(
        model = ~model_fit,
        data = ~standardize,         # <- changed data reference
        xVar = "Temp.Celsius",
        yVar = "Ozone",              # <- new y-variable
        title = "Linear model fit"
    ) {
        coeffs <- coefficients(model)
        ggplot(data) +
            geom_point(aes(.data[[xVar]], .data[[yVar]])) +
            geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
            labs(title = title)
    }
)
```

The updated pipeline now looks as follows.

```{r}
pip
```

```{r, echo = FALSE, eval = nzchar(Sys.getenv("IN_PKGDOWN"))}
library(visNetwork)
do.call(visNetwork, args = c(pip_get_graph(pip), list(height = 100))) |>
    visHierarchicalLayout(direction = "LR")
```

We see that the `model_fit` and `model_plot` steps now use (i.e., depend on) the
standardized data. Let's re-run the pipeline and inspect the output.

```{r}
pip_set_params(pip, params = list(xVar = "Solar.R", yVar = "Wind"))
pip_run(pip)
```

```{r}
pip[["model_fit", "out"]] |> coefficients()
```

```{r, fig.alt = "model-plot", warning = FALSE, message = FALSE}
pip[["model_plot", "out"]]
```


### Removing steps

Let's see the pipeline again.
```{r}
pip
```

When you are trying to remove a step, {pipeflow} by default checks if
the step is used by any other step, and raises an error if removing the
step would violate the integrity of the pipeline.

```{r try-remove-step}
try(pip_remove(pip, "standardize"))
```

To enforce removing a step together with all its downstream
dependencies, you can use the `recursive` argument.

```{r remove-steps-recursively}
pip_remove(pip, "standardize", recursive = TRUE)
```

```{r}
pip
```

Naturally, the last step never has any downstream dependencies, so it
can be removed without any issues.

```{r}
last_step <- tail(pip[["step"]], 1)
pip_remove(pip, last_step)
```

```{r}
pip
```

Replacing steps in a pipeline as shown in this vignette will allow to re-use existing
pipelines and adapt them programmatically to new requirements.
Another way of re-using pipelines is to combine them, which is shown in the
[Combining pipelines](v03-combine-pipelines.html) vignette.

```{r, include = FALSE}
options(old)
```
