---
title: "Working with pipeline views"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  How to filter pipelines with `pip_view()` and the `[` operator, compose
  views, and run only a subset of steps.
vignette: >
  %\VignetteIndexEntry{Working with pipeline views}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

Pipelines can get long, and often you want to focus on a subset of steps:
a topic, a stage, or the steps that produce the outputs you care about.
*Views* are {pipeflow}'s way of working on a subset of steps without
copying anything. A view references the underlying pipeline, so every
operation applied to a view (running it, updating parameters, tagging,
locking, ...) writes through to the original pipeline, restricted to the
steps covered by the view.

This vignette shows how to create and combine views, how to select steps
with the `[` operator, and how to run only part of a pipeline.

## Setup

We use the same example pipeline as in the
`vignette("v04-collect-output")` vignette, tagging steps by topic and
output type.

```{r setup-pipeline}
library(pipeflow)

pip <- pip_new("my-pip") |>
    pip_add("load", \(n = 5) seq_len(n), tags = c("io", "daily")) |>
    pip_add("clean", \(x = ~load) x * 2, tags = c("io", "core")) |>
    pip_add("fit", \(x = ~clean) sum(x), tags = c("model", "core")) |>
    pip_add("report", \(x = ~fit) paste("result:", x), tags = "report")

pip_run(pip, lgr = NULL)
```

## Creating views

`pip_view()` returns a *view* of the pipeline that contains only the
steps matching the given filters:

```{r}
pip_view(pip, tags = "core")
```

Filters can be combined. By default, steps must match *all* filters
(logical AND), while the values within a single filter are treated as
alternatives (OR):

```{r}
pip_view(pip, tags = "core", state = "done")
pip_view(pip, step = c("clean", "fit"))
```

`join = "union"` keeps steps that match *any* filter:

```{r}
pip_view(pip, tags = "report", step = "clean", join = "union")
```

With `fixed = FALSE`, filter values are interpreted as regular
expressions:

```{r}
pip_view(pip, step = "^f", fixed = FALSE)
```

The available filters are `step`, `params`, `state`, `exec`, `tags`, and
`depends`. For example, to find all steps that depend on `load` and are
still `new`:

```{r}
pip_reset(pip)
pip_view(pip, depends = "load", state = "new")
pip_run(pip, lgr = NULL)
```

## Selecting steps with `[`

The extract operator `[` provides a data.table-like way of selecting
steps. It returns a view by default:

```{r}
pip[c("load", "fit")]
pip[1:3]
```

Boolean filters are evaluated in the context of the step table, so the
same columns as in `pip_view()` are available as variables:

```{r}
pip[tags %like% "core"]
pip[step %in% c("clean", "fit") & state == "done"]
```

Negative indices select all steps except the excluded ones:

```{r}
pip[-2]
```

`p[]` returns a copy of the pipeline, and the two-index form extracts
step-table columns as a `data.table`:

```{r}
pip2 <- pip[]
pip[, c("step", "tags")]
pip[c("load", "fit"), "out"]
```

While `[` returns a view by default, `view = FALSE` builds a new,
self-contained pipeline containing the selected steps together with all
their upstream dependencies:

```{r}
pip[c("fit", "report"), view = FALSE]
```

The printed message tells you how many steps were pulled in as upstream
dependencies.

## Composing views

Views can be nested: applying `pip_view()` (or `[`) to a view narrows
the view further.

```{r}
v1 <- pip_view(pip, tags = "io")    # load, clean
v1

v2 <- v1 |> pip_view(tags = "core") # clean only
v2
```

## The `view` meta field

The `view` meta field stores the absolute row indices covered by the
view (`NULL` for a full pipeline). You can restrict a pipeline to a view
by assigning to the meta field, and clear the view again with `NULL`:

```{r}
pip[["view"]]          # NULL — not a view

w <- pip
w[["view"]] <- c("load", "clean")
w

w[["view"]] <- NULL    # back to the full pipeline
w
```

## Running views

Running a view executes the covered steps together with any upstream
dependencies that are not up to date. The run log marks steps that belong
to the view as `[view]` and steps that were pulled in as dependencies as
`[upstream]`:

```{r}
pip_reset(pip)
pip_run(pip_view(pip, step = "report"))
```

Afterwards, the original pipeline is up to date for the covered steps.

## Collecting output from views

Combine views with `pip_collect()` to collect only the outputs you are
interested in:

```{r}
pip |>
    pip_view(tags = "core") |>
    pip_collect()
```

See the `vignette("v04-collect-output")` vignette for more on collecting
output.

```{r, include = FALSE}
options(old)
```
