---
title: "Collecting and grouping output"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to tag, filter, and collect pipeline output.
vignette: >
  %\VignetteIndexEntry{Collecting and grouping output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

Generally speaking, one should keep pipeline steps as simple as possible,
basically following the
principle *"one step, one task"*. While splitting up analysis steps into multiple
functions naturally becomes hard to manage with a growing number of functions,
with {pipeflow}, which manages all function and parameter dependencies for you,
this feels rather easy^[
    The linearly organized set of functions as enforced by the pipeline design
    also helps.
].

Following this principle, most of the pipeline steps will carry intermediate
results while only a few steps contain the final output we are interested in.

This vignette shows how to conveniently tag, collect, and group those final
outputs.

### Tagging steps

Tags let you label steps with meaningful keywords so you can later filter
the pipeline by topic, stage, or output type.
Tags can be set during pipeline creation or later using `pip_set_tags()`.

```{r pipeline-with-output}
library(pipeflow)

pip <- pip_new("my-pip") |>
    pip_add(
        "data",
        function(data = airquality) data
    ) |>
    pip_add("data_prep",
        function(df = ~data) {
            df[, "Temp.Celsius"] <- (df[, "Temp"] - 32) * 5 / 9
            df
        },
        tags = "data" # <-- set 'data' tag
    ) |>
    pip_add(
        "data_summary",
        function(data = ~data_prep,
                 xVar = "Temp.Celsius",
                 yVar = "Ozone") {
            format(summary(data[, c(xVar, yVar)])) |>
                as.data.frame(row.names = NA)
        },
        tags = c("data", "summary") # <-- set 'data' and 'summary' tags
    ) |>
    pip_add(
        "data_plot",
        function(data = ~data_prep,
                 xVar = "Temp.Celsius",
                 yVar = "Ozone") {
            require(ggplot2, quietly = TRUE)
            ggplot(data) +
                geom_point(aes(.data[[xVar]], .data[[yVar]])) +
                labs(title = "Data")
        },
        tags = c("data", "plot") # <-- set 'data' and 'plot' tags
    ) |>
    pip_add(
        "model_fit",
        function(data = ~data_prep,
                 xVar = "Temp.Celsius",
                 yVar = "Ozone") {
            lm(paste(yVar, "~", xVar), data = data)
        },
        tags = c("model", "fit") # <-- set 'model' and 'fit' tags
    ) |>
    pip_add(
        "model_summary",
        function(fit = ~model_fit) {
            summary(fit) |>
                coefficients() |>
                as.data.frame()
        },
        tags = c("model", "summary") # <-- set 'model' and 'summary' tags
    ) |>
    pip_add(
        "model_plot",
        function(model = ~model_fit,
                 data_plot = ~data_plot,
                 xVar = "Temp.Celsius",
                 yVar = "Ozone") {
            coeffs <- coefficients(model)
            data_plot +
                geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
                labs(title = "Linear model fit")
        },
        tags = c("model", "plot") # <-- set 'model' and 'plot' tags
    )
```

We used two families of tags: `"data"`/`"model"` to distinguish the
topic, and `"summary"`/`"plot"`/`"fit"` for the output type.
Whenever tags are defined, they are shown in the pipeline overview
(see rightmost column):

```{r}
pip
```

Before showing how to make use of the tags, let's run the pipeline and inspect
the output individually as we did in the previous vignettes.

```{r}
pip_run(pip)

pip
```

```{r, message = FALSE, warning = FALSE}
pip[["data_plot", "out"]]
```

```{r, message = FALSE, warning = FALSE}
pip[["model_plot", "out"]]
```


### Flat output collection

`pip_collect()` returns all step outputs as a flat named list^[
    Below, we only show the names of the list elements, to keep
    the vignette concise.
].

```{r}
out <- pip_collect(pip)

names(out)
```


### Filtered output using tags

To collect only the output of steps with a specific tag, we use
`pip_view()`, which is {pipeflow}'s general-purpose function for filtering
pipelines, and then call `pip_collect()` on the filtered pipeline view.
For example, to collect only the plots, we can filter by the `plot` tag.

```{r}
pip_view(pip, tags = "plot") # or pip[tags %like% "plot"]
```

```{r grouped-plots, message = FALSE, warning = FALSE}
pip |>
    pip_view(tags = "plot") |>
    pip_collect() |>
    gridExtra::grid.arrange(grobs = _, nrow = 2)
```


### Grouped output via views

Grouped output could be achieved by composing `pip_view()` with
`pip_collect()`. For example, to collect outputs grouped by topic
(`"data"` and `"model"`):

```{r}
grouped <- list(
    data  = pip_view(pip, tags = "data") |> pip_collect(),
    model = pip_view(pip, tags = "model") |> pip_collect()
)

names(grouped[["data"]])
names(grouped[["model"]])
```

However, since this scenario is so common, since version `0.4.0` the
`pip_collect` function natively supports grouping via a `by` parameter,
which allows to simplify the above call as follows.

```{r}
grouped <- pip_collect(pip, by = "tags")[c("data", "model")]

names(grouped[["data"]])
names(grouped[["model"]])
```

In addition, you can now also return the results as a compact table.

```{r}
pip_collect(pip, by = "tags", as.table = TRUE)
```

For more details see `?pip_collect`.

```{r, include = FALSE}
options(old)
```
