---
title: "Collecting and filtering output"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to collect, filter, and group pipeline output.
vignette: >
  %\VignetteIndexEntry{Collecting and filtering output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

Generally speaking, one should keep pipeline steps as simple as possible,
basically following the
principle *"one step, one task"*. Splitting up analysis steps into multiple functions
naturally can be hard to manage, but since {pipeflow} manages all function and parameter dependencies for you, this is not a problem.
Following this principle, usually a lot of pipeline steps will carry intermediate results
and only a few steps will contain the final output we are interested in.
This vignette shows how to conveniently tag, collect, and possibly group those final
outputs.

### Group and tag definitions

The {pipeflow} package is developed with the goal in mind to provide a fast,
lightweight, and flexible framework that does not force users into a specific
workflow.
However, two features for managing pipeline output seemed to appear so
frequently in practice that they found their way as core features into the
package:

* groups to define related output steps, and
* tags to mark steps and enable to filter them for output collection.

Both **group** and **tag** properties can be already set during the pipeline
creation. Let's illustrate this with an example.

```{r pipeline-with-output}
library(pipeflow)

pip <- pip_new("my-pip") |>
    pip_add(
        "data",
        function(data = airquality) data
    ) |>
    pip_add("data_prep",
        function(data = ~data) {
            replace(data, "Temp.Celsius", (data[, "Temp"] - 32) * 5 / 9)
        },
        group = "Data", # <- set 'Data' group
        tags = "data" # <- set 'data' tag
    ) |>
    pip_add(
        "data_summary",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            format(summary(data[, c(xVar, yVar)])) |>
                as.data.frame(row.names = NA)
        },
        group = "Data", # <- set 'Data' group
        tags = c("data", "summary") # <- set 'data' and 'summary' tags
    ) |>
    pip_add(
        "data_plot",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            require(ggplot2, quietly = TRUE)
            ggplot(data) +
                geom_point(aes(.data[[xVar]], .data[[yVar]])) +
                labs(title = "Data")
        },
        group = "Data", # <- set 'Data' group
        tags = c("data", "plot") # <- set 'data' and 'plot' tags
    ) |>
    pip_add(
        "model_fit",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            lm(paste(yVar, "~", xVar), data = data)
        },
        group = "Model", # <- set 'Model' group
        tags = c("model", "fit") # <- set 'model' and 'fit' tags
    ) |>
    pip_add(
        "model_summary",
        function(fit = ~model_fit) {
            summary(fit) |>
                coefficients() |>
                as.data.frame()
        },
        group = "Model", # <- set 'Model' group
        tags = c("model", "summary") # <- set 'model' and 'summary' tags
    ) |>
    pip_add(
        "model_plot",
        function(
            model = ~model_fit,
            data_plot = ~data_plot,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            coeffs <- coefficients(model)
            data_plot +
                geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
                labs(title = "Linear model fit")
        },
        group = "Model", # <- set 'Model' group
        tags = c("model", "plot") # <- set 'model' and 'plot' tags
    )
```

Looking at the pipeline, we can see that in contrast to the examples
of the previous vignettes, two new columns `group` and `tags` have been
added to the overview.

```{r}
pip
```

Before showing how to make use of the defined groups and tags, let's run the
pipeline and inspect the output inidividually as we did in the previous
vignettes.

```{r}
pip_run(pip)

pip
```

```{r, message = FALSE, warning = FALSE}
pip[["data_plot", "out"]]
```

```{r, message = FALSE, warning = FALSE}
pip[["model_plot", "out"]]
```


### Grouped output

The default {pipeflow} way of gobbling up all output is to call
`pip_collect_out()` on the pipeline, which returns a list with all
the output.

```{r}
out <- pip_collect_out(pip)

str(out, max.level = 1)
```

But, as we can see, the output has been grouped into sublists defined by the
`group` properties set earlier. The `Data` and `Model` sublists contain the
output of the respective steps.

```{r}
names(out[["Data"]])

names(out[["Model"]])
```

To ignore the defined groups and collect all output step-wise into a single
flat list, we set `grouped = FALSE` when calling `pip_collect_out()`.

```{r}
out <- pip_collect_out(pip, grouped = FALSE)

names(out)
```


### Filtered output using tags

To collect only the output of steps with a specific tag, we use
`pip_view()`, which is {pipeflow}'s general-purpose function for filtering
pipelines and then call `pip_collect_out()` on the filtered pipeline.
To collect only the plots, for example, we can filter by the `plot` tag.

```{r}
pip_view(pip, tags = "plot")
```

```{r, message = FALSE, warning = FALSE}
pip |>
    pip_view(tags = "plot") |>
    pip_collect_out() |>
    gridExtra::grid.arrange(grobs = _, nrow = 2)
```

Next, we collect all summaries.

```{r, message = FALSE, warning = FALSE}
summaries <- pip |>
    pip_view(tags = "summary") |>
    pip_collect_out()
```

```{r}
summaries[["Data"]]
```

```{r}
summaries[["Model"]]
```

### More on views

To make the example a bit more interesting, we first update some parameters.

```{r}
pip_set_params(pip, params = list(xVar = "Solar.R", yVar = "Wind"))

pip
```

{pipeflow} views provide a variety of filtering options. In the previous
section, the filtering was done based on tags, but you can also filter
based on other properties, for example, all steps that depend on the
`model_fit` step:

```{r}
pip |> pip_view(filter = list(depends = "model_fit", state = "outdated"))
```

or using regex-based filtering, for example, to filter all outdated steps
starting with `data`:

```{r}
pip |>
    pip_view(filter = list(step = "^data", state = "outdated"), fixed = FALSE)
```

Views can also be chained together:

```{r}
v <- pip |> pip_view(filter = list(state = "outdated"))
v

v2 <- v |> pip_view(tags = "plot")
v2
```

Last but not least, views can be run as pipelines themselves, which allows to
conveniently re-run only the filtered steps, while {pipeflow} ensures that
any upstream dependencies are run first if needed.

```{r}
v2 |> pip_run()
```

Having a closer look at the run log, you'll see which steps were re-run as
part of the `[view]` and which were re-run as `[upstream]` dependencies.
Since all views work by reference on the given pipeline, the original pipeline
is now up-to-date for the filtered steps.

```{r}
pip
```

```{r, include = FALSE}
options(old)
```
