---
title: "Collecting and filtering output"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to tag, filter, and collect pipeline output.
vignette: >
  %\VignetteIndexEntry{Collecting and filtering output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

Generally speaking, one should keep pipeline steps as simple as possible,
basically following the
principle *"one step, one task"*. While splitting up analysis steps into multiple
functions naturally becomes hard to manage with a growing number of functions,
with {pipeflow}, which manages all function and parameter dependencies for you,
this feels rather easy.

Following this principle, most of the pipeline steps will carry intermediate
results while only a few steps contain the final output we are interested in.

This vignette shows how to conveniently tag and collect those final outputs
as well as how to filter pipelines via {pipeflow}s view functionality.

### Tagging steps

Tags let you label steps with meaningful keywords so you can later filter
the pipeline by topic, stage, or output type.
Tags can be set during pipeline creation or later using `pip_set_tags()`.

```{r pipeline-with-output}
library(pipeflow)

pip <- pip_new("my-pip") |>
    pip_add(
        "data",
        function(data = airquality) data
    ) |>
    pip_add("data_prep",
        function(data = ~data) {
            replace(data, "Temp.Celsius", (data[, "Temp"] - 32) * 5 / 9)
        },
        tags = "data"                   # <-- set 'data' tag
    ) |>
    pip_add(
        "data_summary",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            format(summary(data[, c(xVar, yVar)])) |>
                as.data.frame(row.names = NA)
        },
        tags = c("data", "summary")     # <-- set 'data' and 'summary' tags
    ) |>
    pip_add(
        "data_plot",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            require(ggplot2, quietly = TRUE)
            ggplot(data) +
                geom_point(aes(.data[[xVar]], .data[[yVar]])) +
                labs(title = "Data")
        },
        tags = c("data", "plot")        # <-- set 'data' and 'plot' tags
    ) |>
    pip_add(
        "model_fit",
        function(
            data = ~data_prep,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            lm(paste(yVar, "~", xVar), data = data)
        },
        tags = c("model", "fit")        # <-- set 'model' and 'fit' tags
    ) |>
    pip_add(
        "model_summary",
        function(fit = ~model_fit) {
            summary(fit) |>
                coefficients() |>
                as.data.frame()
        },
        tags = c("model", "summary")    # <-- set 'model' and 'summary' tags
    ) |>
    pip_add(
        "model_plot",
        function(
            model = ~model_fit,
            data_plot = ~data_plot,
            xVar = "Temp.Celsius",
            yVar = "Ozone"
        ) {
            coeffs <- coefficients(model)
            data_plot +
                geom_abline(intercept = coeffs[1], slope = coeffs[2]) +
                labs(title = "Linear model fit")
        },
        tags = c("model", "plot")       # <-- set 'model' and 'plot' tags
    )
```

We used two families of tags: `"data"`/`"model"` to distinguish the
topic, and `"summary"`/`"plot"`/`"fit"` for the output type.
Whenever tags are defined, they are shown in the pipeline overview
(see rightmost column):

```{r}
pip
```

Before showing how to make use of the tags, let's run the pipeline and inspect
the output individually as we did in the previous vignettes.

```{r}
pip_run(pip)

pip
```

```{r, message = FALSE, warning = FALSE}
pip[["data_plot", "out"]]
```

```{r, message = FALSE, warning = FALSE}
pip[["model_plot", "out"]]
```


### Flat output collection

`pip_collect_out()` returns all step outputs as a flat named list^[
    Below, we only show the names of the list elements, to keep
    the vignette concise.
].

```{r}
out <- pip_collect_out(pip)

names(out)
```


### Filtered output using tags

To collect only the output of steps with a specific tag, we use
`pip_view()`, which is {pipeflow}'s general-purpose function for filtering
pipelines, and then call `pip_collect_out()` on the filtered pipeline view.
For example, to collect only the plots, we can filter by the `plot` tag.

```{r}
pip_view(pip, tags = "plot")
```

```{r grouped-plots, message = FALSE, warning = FALSE}
pip |>
    pip_view(tags = "plot") |>
    pip_collect_out() |>
    gridExtra::grid.arrange(grobs = _, nrow = 2)
```


### Grouped output via views

Grouped output is achieved by composing `pip_view()` with
`pip_collect_out()`. For example, to collect outputs grouped by topic
(`"data"` and `"model"`):

```{r}
grouped <- list(
    data  = pip_view(pip, tags = "data")  |> pip_collect_out(),
    model = pip_view(pip, tags = "model") |> pip_collect_out()
)

names(grouped[["data"]])
names(grouped[["model"]])
```


### More on views

To make the example a bit more interesting, we first update some parameters.

```{r}
pip_set_params(pip, params = list(xVar = "Solar.R", yVar = "Wind"))

pip
```

{pipeflow} views provide a variety of filtering options. In the previous
section, the filtering was done based on tags, but you can also filter
based on other properties, for example, to select all `outdated` steps
that depend on the `model_fit` step:

```{r}
pip |> pip_view(filter = list(depends = "model_fit", state = "outdated"))
```

or using regex-based filtering to filter all outdated steps
starting with `data`:

```{r}
pip |>
    pip_view(filter = list(step = "^data", state = "outdated"), fixed = FALSE)
```

Views can also be chained together:

```{r}
v <- pip |> pip_view(filter = list(state = "outdated"))
v

v2 <- v |> pip_view(tags = "plot")
v2
```

Last but not least, views can be run as pipelines themselves, which allows to
conveniently re-run only the filtered steps, while {pipeflow} ensures that
any upstream dependencies are run first if needed.

```{r}
v2 |> pip_run()
```

Having a closer look at the run log, you'll see which steps were re-run as
part of the `[view]` and which were re-run as `[upstream]` dependencies.
Since all views work by reference on the given pipeline, the original pipeline
is now up-to-date for the filtered steps.

```{r}
pip
```

```{r, include = FALSE}
options(old)
```
