---
title: "Nested pipelines"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how to embed pipelines within pipeline steps and how to mark their
  parameters so that changing them re-executes the outer step.
vignette: >
  %\VignetteIndexEntry{Nested pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```


A pipeline step can contain another pipeline: instead of computing a result
directly, the step builds an inner pipeline and returns it. This allows to
reuse a standard analysis pipeline inside a larger workflow.

### Inner pipeline

We start with the same pipeline as in the previous
["Split, map, and reduce"](v05-split-map-reduce.html) vignette,
fitting a linear model and returning its coefficients. This will
serve as our inner pipeline.

```{r inner-pipeline}
library(pipeflow)

inner <- pip_new("coefficients") |>
    pip_add("data", \(data = NULL) data) |>
    pip_add(
        "fit",
        \(data = ~data, xVar = "x", yVar = "y") {
            lm(paste(yVar, "~", xVar), data = data)
        }
    ) |>
    pip_add("coefs", \(fit = ~fit) coefficients(fit))

inner
```


### Outer pipeline

The outer pipeline splits the data into subsets and derives the
model coefficients by running the inner pipeline for each split.

```{r outer-pipeline}
# Helper to run inner pipeline
run_inner_pip <- function(pip, name, data) {
    pip$name <- sprintf("coefs for *%s*", name)

    # Set data subset for inner pipeline and run it
    pip_set_params(pip, list(data = data)) |> pip_run()

    pip[["coefs", "out"]]
}

outer <- pip_new("full analysis") |>
    pip_add("data", \(data = NULL) data) |>
    pip_add(
        "split_data", \(data = ~data, byVar = "by") {
            split(data, f = data[[byVar]])
        }
    ) |>
    pip_add(
        "inner_run",
        \(dataList = ~split_data, xVar = "x", yVar = "y") {
            p <- pip_clone(inner)

            # Forward parameters to inner
            pip_set_params(p, list(xVar = xVar, yVar = yVar))

            Map(
                f = run_inner_pip,
                name = names(dataList),
                data = dataList,
                MoreArgs = list(pip = p)
            )
        }
    ) |>
    pip_add(
        "combine",
        \(coefs = ~inner_run) as.data.frame(do.call(rbind, coefs))
    )

outer
```

Note the `inner_run` step: its parameters `xVar` and `yVar` are forwarded
to the inner pipeline.

### Run nested pipeline
Let's now set the analysis parameters and run the full pipeline:

```{r}
outer |>
    pip_set_params(
        list(
            data = iris,
            xVar = "Sepal.Length",
            yVar = "Sepal.Width",
            byVar = "Species"
        )
    ) |>
    pip_run()
```

The output of the `inner_run` step is a list of coefficient vectors, one
for each species,

```{r}
outer[["inner_run", "out"]]
```

and the `combine` step returns the expected combined table.

```{r}
outer[["combine", "out"]]
```


Now suppose we want to change one of the model settings, say use
`Petal.Length` instead of `Sepal.Length` as the predictor.

```{r}
pip_set_params(outer, params = list(xVar = "Petal.Length"))

outer
```

Since the `xVar` parameter is part of the `inner_run` step's function arguments,
the `inner_run` step's state (and its downstream dependencies) correctly has
now been marked as "outdated".

### Forward inner pipeline parameters programmatically

While forwarding the parameters *manually* to the inner pipeline is
straight-forward in our toy example, trying to *manually* synchronize
real-world parameter sets between the outer and inner pipeline  quickly
becomes a unfeasible and a source for bugs that are hard to detect.

For this reason, in practice, the following pattern should be used, which
basically just forwards the combined set of *all* existing parameters.
To do this, we replace the `inner_run` step as follows:

```{r replace-inner-run}
outer |> pip_replace(
    "inner_run",
    \(dataList = ~split_data, ...) {
        p <- pip_clone(inner)

        # Forward all parameters (from outer and inner)
        all_params <- .self$get_params()
        pip_set_params(p, all_params)

        Map(
            f = run_inner_pip,
            name = names(dataList),
            data = dataList,
            MoreArgs = list(pip = p)
        )
    },
    params = pip_get_params(inner) # <--- default parameters of inner pipeline
)

outer
```

In this version, the `inner_run` step no longer declares `xVar` and `yVar` as
its own arguments. Instead, its parameters are seeded with the inner
pipeline's parameters via `params = pip_get_params(inner)`, and the step
forwards the combined parameter set at run time. Three aspects are worth
spelling out:

* **`.self`** refers to the pipeline that is currently being run and is
  available inside *every* step function without having to declare it. It
  exposes the usual pipeline interface, so `.self$get_params()` returns the
  current unbound parameters of the outer pipeline, and things like
  `.self$name` or `.self$run()` work as expected. For more examples on this
  mechanism see the
  [self-modifying pipelines](v06-self-modify-pipeline.html) vignette.
* **`params = pip_get_params(inner)`** registers the inner pipeline's unbound
  parameters (`data`, `xVar`, `yVar`) as parameters of the `inner_run` step.
  They therefore show up in the step's `params` column, become part of the
  outer pipeline's parameter set (and can be updated with
  `pip_set_params(outer, ...)`), and are passed to the step function when it
  runs.
* **Precedence over function arguments**: when `params` overlaps with the
  arguments of the step function, the function's default values take
  precedence. This is why the replacement function above declares only
  `dataList` and `...` — had it declared e.g. `xVar = "x"`, that default would
  win over the value coming from `params`, and the forwarding would silently
  be ignored. As another example,
  `pip_add("s", \(x = 99, ...) x, params = list(x = 1))` stores `x = 99`,
   not `x = 1`.

Let's re-run the full pipeline.

```{r re-run-full-pipe}
outer |>
    pip_set_params(
        list(
            data = iris,
            xVar = "Sepal.Length",
            yVar = "Sepal.Width",
            byVar = "Species"
        )
    ) |>
    pip_run()
```

Note that the warning in the log above is expected and harmless. Basically,
`.self$get_params()` returns the outer pipeline's *entire* parameter set,
which includes `byVar` (from the `split_data` step) while the inner
pipeline only defines `data`, `xVar`, and `yVar`. Since `pip_set_params()`
reports any parameters that are not defined in the target and simply leaves
them unset, the inner pipeline still receives all parameters it knows and the
result is unaffected.

If you need to omit the warning (e.g. in production code), just suppress it:

```r
suppressWarnings(pip_set_params(p, all_params))
```

Alternatively, you could first restrict the forwarded parameters to those the
inner pipeline actually knows. That has the same effect but adds code, and
since forwarding the full parameter set is the whole point of this pattern,
there is no need for it.

Again, changing one of the inner parameters will correctly outdate the
`inner_run` step plus downstream dependencies.

```{r}
pip_set_params(outer, params = list(xVar = "Petal.Length"))

outer
```

With the above pattern, you can now change both the inner and outer pipeline,
adding and/or removing any steps or parameters, without having to worry
about parameter synchronization.

```{r, include = FALSE}
options(old)
```
