---
title: "Self-modifying pipelines"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Shows how you can setup pipelines to modify themselves at runtime, which,
  for example, allows for changing pipeline parameters based on intermediate
  results or even dynamically modify the pipeline's own structure during
  a pipeline run.

vignette: >
  %\VignetteIndexEntry{Self-modifying pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
require(pipeflow)

knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

The {pipeflow} package aims to offer a lean and intuitive interface that
enables new users to get started quickly without having to learn a lot of
new concepts and functions.

At the same time, it was also designed to provide easy access to the
underlying data structures to allow advanced users to modify the
pipeline basically in any way they want.
As such, pipelines not only can be modified before but also during
execution.
This opens up a wide range of possibilities, for example,
to change parameters based on intermediate results or even to modify the
pipeline structure itself during a pipeline run.
In this vignette, we will present some basic examples of how this can be done.

### The pipeline

Let's first define a pipeline, which fits a linear model, checks it's
residuals for normality using the Shapiro-Wilk test, and plots the
residuals.

```{r}
library(pipeflow)

pip <- pip_new("residual-analysis") |>
    pip_add("data", \(data = NULL) data) |>
    pip_add("fit", \(data = ~data, xVar = "x", yVar = "y") {
        lm(paste(yVar, "~", xVar), data = data)
    }) |>
    pip_add(
        "shapiro",
        \(fit = ~fit) {
            shapiro.test(residuals(fit))$p.value
        }
    ) |>
    pip_add(
        "plot",
        \(fit = ~fit, pointColor = "black") {
            require(ggplot2, quietly = TRUE)
            data <- data.frame(fitted = predict(fit), resids = residuals(fit))
            ggplot(data, aes(x = fitted, y = resids)) +
                geom_point(shape = 21, color = pointColor) +
                geom_hline(yintercept = 0, linetype = "dashed") +
                theme_minimal()
        }
    )
```

By now, you should be familar with the pipeline's overview.

```{r}
pip
```

#### Internal pipeline structure

Before we move on, let's briefly inspect the internal structure of a
pipeline object.

```{r}
str(pip)
```

There is the `name` of the pipeline, which is just a character string,
as well as a `view` entry, which is currently undefined.

The most "interesting" part is the `pipenv`: an environment that holds the
pipeline's actual state. It is shared by the pipeline and all of its views,
which is why operations on a view write through to the underlying pipeline.

```{r}
ls(pip$pipenv)
```

Listing the `pipenv`^[
    By default, `ls()` hides names starting with a dot. The internal
    entries, which become visible via `ls(pip$pipenv, all.names = TRUE)`,
    are not considered in this vignette.
] reveals the `data`entry, which contains the pipeline's step table as a
`data.table` with one row per step:

```{r}
pip$pipenv$data
```

```{r}
str(pip$pipenv$data, max.level = 1)
```

Its columns describe a step and its last result:

* `step`, `fun`, `params`, `out`, `state`, `tags`, `locked`, `exec`, and
  `time`: the step's name, function, parameters, last output, state, tags,
  lock flag, execution mode, and time stamp,
* `depends`: the names of the steps this step depends on,
* `unbound`: the names of the step's adjustable (non-dependency) parameters,
* `nodeId`: the step's node id in the internal dependency graph.

The last three columns are derived from the step definition. {pipeflow}
recomputes them, and they cannot be assigned to directly.

Instead of reaching into `pip$pipenv$data` yourself, you can use
`pip_data(pip)` to retrieve the step table, or the `[` and `[[` operators and
`print(pip)` to inspect selected parts. This internal structure is what the
following examples modify at runtime; note that changing the step table
directly can easily invalidate the internal dependency graph, so for
structural changes prefer the dedicated functions such as `pip_add()`,
`pip_replace()`, and `pip_remove()`.


### Changing parameters at runtime

First, we set some data and parameters and run the pipeline as usual.

```{r, fig.alt = "residual-plot"}
pip |> pip_set_params(list(data = airquality, xVar = "Ozone", yVar = "Temp"))

pip_run(pip)

pip[["plot", "out"]]
```

Now let's imagine, we want to change the color of the points in the plot
depending on the Shapiro-Wilk test result. The obvious way to do this
would be to change the `plot` step by passing the test result to the
`plot` step function and change the color there.

However, here we are interested in a solution that keeps the `plot`
function unchanged.
Naively, we could run the pipeline a second time as follows:

```{r, fig.alt = "residual-plot"}
if (pip[["shapiro", "out"]] < 0.05) {
    pip |>
        pip_set_params(list(pointColor = "red")) |>
        pip_run()
}

pip[["plot", "out"]]
```

This solution is not ideal, as it requires to run additional code around
the pipeline. We rather want to set the parameter from within the
pipeline during execution.

For this, the pipeline by default provides a self-reference in the
form of the `.self` object that is automatically made available in
any step function.
With this in mind, we can update the `shapiro`
step to change the `pointColor` parameter on the fly.

```{r}
pip |> pip_replace(
    "shapiro",
    \(fit = ~fit) {
        residuals <- residuals(fit)
        p <- shapiro.test(residuals)$p.value

        if (p < 0.05) {
            .self$set_params(list(pointColor = "blue"))
            # or: .pip_set_params(.self, list(pointColor = "blue"))
        }

        p
    }
)
```

Now we can run the pipeline again and see that the color of the points
has changed.

```{r, fig.alt = "residual-plot2"}
pip_run(pip)

pip[["plot", "out"]]
```

As we will show in the next section, modications of the pipeline at
runtime are not limited to changing parameters but can also be used
to modify the very own pipeline structure.

### Changing the pipeline structure at runtime

Subsequently, the pipeline steps will be comprised of very basic
functions to keep the focus on the pipeline structure and how it is
modified.

```{r}
pip <- pip_new("my-pipeline") |>
    pip_add("init", \(xInit = 0) xInit) |>
    pip_add("f1", \(x = ~init) x + 1) |>
    pip_add("f2", \(x = ~f1) x + 2) |>
    pip_add("f3", \(x = ~f2) x + 3)
```

This pipeline just adds 1, 2, and 3 to the initial value, respectively.

```{r}
pip_run(pip)

pip
```

The `out` column in the table shows the output of each step.
Now let's modify step `f2` that in turn will modify `f3` at
runtime based on the interim result passed into `f2`.

#### Modify steps

```{r}
pip |> pip_replace(
    "f2",
    \(x = ~f1) {
        if (x > 10) {
            .self$replace("f3", \(x = ~f1) x * 3)
            # or: pip_replace(.self, "f3", \(x = ~f1) x * 3)
            return(x / 2)
        }
        x + 2
    }
)
```

Basically, step `f2` now checks if the input is greater than 10,
and if so, it replaces step `f3` with a new step now referencing `f1`
that multiplies the input passed from `f1` by 3 and returns half of
the input.

To see this, let's try it with an input of 15.

```{r}
pip |>
    pip_set_params(list(xInit = 15)) |>
    pip_run()

pip
```

We see that both the output of the pipeline and the dependencies
of the last step have changed. Let's confirm by inspecting the function
of the last step.

```{r}
pip[["f3", "fun"]]
```

#### Insert and remove steps

For our last example, we get even more hacky and instead of just replacing,
we will go a bit further to insert and remove steps.
The pipeline definition is as follows:

```{r}
pip <- pip_new("my-pipeline") |>
    pip_add("init", \(xInit = 0) xInit) |>
    pip_add("f1", \(x = ~init) x + 1) |>
    pip_add(
        "f2",
        \(x = ~f1) {
            if (x > 10) {
                .self |>
                    pip_add("f2a", \(x = ~f1) x + 21, after = "f1") |>
                    pip_add("f2b", \(x = ~f2a) x + 22, after = "f2a") |>
                    pip_replace("f3", \(x = ~f2b) x + 30) |>
                    pip_remove("f2")
            }
            x + 2
        }
    ) |>
    pip_add("f3", \(x = ~f2) x + 3)
```


Basically, if the input is greater than 10, we insert two new steps
`f2a` and `f2b` after `f1`, remove `f2`, and replace `f3` with a new
step that adds 30 to the input. Let's first run with the initial
value of 0 to see the original output.

```{r}
pip_run(pip)

pip
```

Next, we set the initial value to 11 to trigger the changes.

```{r}
pip |>
    pip_set_params(list(xInit = 11)) |>
    pip_run()

pip
```

While the structure has changed as expected, some steps were not
yet run. In fact, since originally step `f3`came after `f2`,
and in contrast to what the log is showing, instead of step `f3`,
actually the new step `f2b` was run^[
    Compare the state of `f2b`, which is set to `done`.
]
, albeit with x = NULL as input.

So to have the true results, we need to re-init the parameter
and need to re-run the pipeline.

```{r}
pip |>
    pip_set_params(list(xInit = 11)) |>
    pip_run()

pip
```

Now the output of all steps is as expected. If we want to use
{pipeflow} in production, obviously, having to re-run the pipeline
and temporarily showing a wrong log is not ideal. Ideally, the
pipeline run would be aborted right after all changes were
done in `f2` and the pipeline re-run automatically from the beginning.
Also, this process potentially should be repeated recursively until
the structure does not change anymore.

Luckily, with some minimal changes, this behaviour can be achieved
with {pipeflow}.
First, for any step where you want to restart the pipeline run, you
need to call `.self$restart()`, so we adapt the `f2` function as
follows:

```{r}
pip <- pip_new("my-pipeline") |>
    pip_add("init", \(xInit = 0) xInit) |>
    pip_add("f1", \(x = ~init) x + 1) |>
    pip_add(
        "f2",
        \(x = ~f1) {
            if (x > 10) {
                .self |>
                    pip_add("f2a", \(x = ~f1) x + 21, after = "f1") |>
                    pip_add("f2b", \(x = ~f2a) x + 22, after = "f2a") |>
                    pip_replace("f3", \(x = ~f2b) x + 30) |>
                    pip_remove("f2")
                .self$restart() # <-- restart the run
            }
            x + 2
        }
    ) |>
    pip_add("f3", \(x = ~f2) x + 3)
```

Second, you just run the pipeline as usual.

```{r}
pip |>
    pip_set_params(list(xInit = 11)) |>
    pip_run()
```

As you can see, the run is now automatically aborted right after
the pipeline was modified and, since the step called
`.self$restart()`, the pipeline is also restarted automatically
based on the new structure.
As a result, the log now is fully aligned with the performed
pipeline run.

Looking at the final pipeline overview, we see that the output matches
the expected output of the modified pipeline.

```{r}
pip
```

Of course, this was just a silly example to show some possibilities,
but I have made use of this feature already in various projects and
may present one of them in a more sophisticated example in the future.

Lastly note that since you have full access to the pipeline object,
of course, you can get even more hacky, but be aware that some
additional operations are done under the hood when steps are added
or removed. It is therefore not recommended to "manually" manipulate
the internal data.table object in terms of removing or adding rows,
or changing important columns such as `depends` or `nodeId` as
this immediately would invalidate the internal consistency of the
dependency graph.

On the other hand, changing entries in columns such as
`tags`, `time`, `state` or `output` is generally not critical.
If in doubt, just try and see what works.
