---
title: "pipeflow vs targets benchmark"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  md_document:
    variant: gfm
    preserve_yaml: true
params:
    nrep: 30
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

nrep <- params$nrep
old <- options(width = 100L)
```

```{r libs, include = FALSE}
devtools::load_all()
library(microbenchmark)
library(targets)
suppressPackageStartupMessages({
    library(readr)
    library(dplyr)
    library(ggplot2)
})
```


Package: pipeflow `r packageVersion("pipeflow")`, targets `r packageVersion("targets")`

## 1. Walkthrough example

Based on https://books.ropensci.org/targets/walkthrough.html

```{r helpers}
# General helpers
create_data_csv <- function(data = airquality, file = "data.csv") {
    utils::write.csv(data, file)
}
elapsed_time <- function(expr) {
    system.time(expr, gcFirst = FALSE)[["elapsed"]]
}


# Shared helpers used by both pipeflow and targets for step definitions
get_data <- function(file) {
    utils::read.csv(file) |> stats::na.omit()
}

fit_model <- function(data) {
    lm(Ozone ~ Temp, data) |> coefficients()
}

plot_model <- function(model, data) {
    ggplot(data) +
        geom_point(aes(x = Temp, y = Ozone)) +
        geom_abline(intercept = model[1], slope = model[2])
}
```


### pipeflow pipeline

```{r pipeflow-walkthrough-run}
tar_dir({
    # Setup data and pipeline
    create_data_csv(file = "data.csv")
    p <- pip_new("walkthrough") |>
        pip_add("data",  \(file = "data.csv") get_data(file)) |>
        pip_add("model", \(data = ~data) fit_model(data)) |>
        pip_add("plot",  \(model = ~model, data = ~data) plot_model(model, data))


    message("\nProof of principle full run (no skips)")
    pip_run(p)
    p_r <- replicate(nrep, elapsed_time(pip_run(p, lgr = NULL, force = TRUE)))

    message("\nProof of principle skipped run")
    pip_run(p)
    p_s <- replicate(nrep, elapsed_time(pip_run(p, lgr = NULL)))
})
```

### targets pipeline

```{r targets-walkthrough-run}
tar_make_here <- function(reporter = "silent") {
    tar_make(callr_function = NULL, reporter = reporter)
}

tar_dir({
    # Setup data and pipeline
    create_data_csv(file = "data.csv")
    tar_script({
        list(
            tar_target(file, "data.csv", format = "file"),
            tar_target(data, get_data(file)),
            tar_target(model, fit_model(data)),
            tar_target(plot, plot_model(model, data))
        )
    }, ask = FALSE)

    message("\nProof of principle full run (no skips)")
    tar_make_here(reporter = "timestamp")
    tar_r <- replicate(nrep, {
        tar_destroy(ask = FALSE)
        elapsed_time(tar_make_here(reporter = "silent"))
    })

    message("\nProof of principle skipped run")
    tar_make_here(reporter = "timestamp")
    tar_s <- replicate(nrep, {
        elapsed_time(tar_make_here(reporter = "silent"))
    })
})
```

### Runtimes

```{r walkthrough-runtime-plot, echo = FALSE, fig.width = 7, fig.height = 4}
title <- sprintf("Walkthrough (4 steps) — %d iterations", nrep)

data <- data.frame(
    "Package" = c("pipeflow", "targets") |> rep(each = 2 * nrep),
    "Operation" = c("full run", "skipped run") |> rep(each = nrep),
    "Time" = c(p_r, p_s, tar_r, tar_s) * 1000
)

ggplot(data, aes(x = Operation, y = Time, fill = Package)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    labs(y = "Time (ms)", x = NULL, title = title) +
    theme_bw()
```

## 2. Long linear pipeline (N sequential steps)

Each step depends on the output of the previous step: `s0 -> s1 -> s2 -> ... -> sN`.
We don't care about the actual computations here, just the overhead of managing
the pipeline structure and skipping logic as the number of steps increases.

### pipeflow pipeline

```{r linear-pipeflow}
create_linear_pip <- function(n) {
    pip <- pip_new("linear") |> pip_add("s0", \(init = 0) init)

    # Add steps 1 to n where each step links to the prior step
    for (i in seq_len(n)) {
        pip_add(pip, step = paste0("s", i), \(x = ~ -1) x + 1)
    }
    pip
}

# Verify
p <- create_linear_pip(3)
pip_run(p)
stopifnot(p[["s3", "out"]] == 3)
```

```{r linear-targets}
create_linear_tar <- function(n) {
    init <- tar_target(s0, 0)
    rest <- lapply(
        seq_len(n),
        FUN = \(i) tar_target_raw(
            sprintf("s%d", i),
            call("+", as.symbol(sprintf("s%d", i - 1)), 1)
        )
    )
    c(list(init), rest)
}

# Verify
tar_dir({
    tar_script(create_linear_tar(3), ask = FALSE)
    tar_make_here(reporter = "timestamp")
    stopifnot(tar_read(s3) == 3)
})
```

### Benchmark

```{r linear-benchmark, echo = FALSE}
pipeline_sizes <- c(16, 32, 64, 128)

# ---- pipeflow: build once, time only forced re-runs ----
pips_pf <- lapply(pipeline_sizes, create_linear_pip) |>
    stats::setNames(paste0("pf_n=", pipeline_sizes))

pf_times <- lapply(pips_pf, function(p) {
    replicate(nrep, elapsed_time(pip_run(p, lgr = NULL, force = TRUE)))
})


# ---- targets: fresh tar_dir each iteration, only tar_make is timed ----
tar_times <- lapply(pipeline_sizes, function(n) {
    replicate(nrep, {
        tar_dir({
            tar_script(create_linear_tar(n), ask = FALSE)
            elapsed_time(tar_make(callr_function = NULL, reporter = "silent"))
        })
    })
})
```

```{r linear-plot, echo = FALSE, fig.width = 7, fig.height = 4}

# Reshape into a single data.frame for plotting
pf_df <- do.call(rbind, lapply(seq_along(pf_times), function(i) {
    data.frame(expr = names(pf_times)[i], time = pf_times[[i]])
}))
tar_df <- do.call(rbind, lapply(seq_along(tar_times), function(i) {
    data.frame(expr = paste0("tar n=", pipeline_sizes[i]), time = tar_times[[i]])
}))

d_pf  <- cbind(pf_df,  package = "pipeflow")
d_tar <- cbind(tar_df, package = "targets")

d_all <- rbind(d_pf, d_tar)
d_all$n <- as.numeric(gsub("[^0-9]", "", d_all$expr))

title <- sprintf("Linear pipeline — %d iterations", nrep)
ggplot(d_all, aes(x = factor(n), y = time * 1000, fill = package)) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    labs(y = "Time (ms)", x = "Number of steps", title = title) +
    theme_bw()
```

## 3. DAG with branching (fan-out: N parallel branches → 1 sink)

Source feeds N parallel branches, all converging on a single sink step.

```{r dag-helpers}
# Shared helpers for the fan-out DAG — used by both pipeflow and targets
dag_source <- function() 1
dag_branch <- function(x) x + 1
dag_sink   <- function(...) sum(...)
```

```{r dag-pipeflow}
make_branch_pip <- function(n) {
  pip <- pip_new("dag") |> pip_add("source", dag_source)
  for (i in seq_len(n))
    pip_add(pip, paste0("b", i), \(x = ~source) dag_branch(x))

  sink_args <- paste(sprintf("x%s = ~b%s", seq_len(n), seq_len(n)), collapse = ", ")
  sink_call <- paste(paste0("x", seq_len(n)), collapse = ", ")
  eval(parse(text = sprintf(
    "pip_add(pip, 'sink', function(%s) { dag_sink(%s) })", sink_args, sink_call
  )))
  invisible(pip)
}

p4 <- make_branch_pip(4); pip_run(p4)
stopifnot(p4[["sink", "out"]] == 8)
```

```{r dag-targets}
make_branch_tar <- function(n) {
  source <- tar_target(source, dag_source())
  branches <- lapply(
    seq_len(n),
    FUN = \(i) tar_target_raw(
      sprintf("b%d", i),
      call("dag_branch", as.symbol("source"))
    )
  )
  sink <- tar_target_raw(
    "sink",
    as.call(c(as.symbol("dag_sink"), lapply(paste0("b", seq_len(n)), as.symbol)))
  )
  c(list(source), branches, list(sink))
}

# Verify
tar_dir({
  tar_script(make_branch_tar(4), ask = FALSE)
  tar_make_here()
  stopifnot(tar_read(sink) == 8)
})
```

### Benchmark

```{r dag-benchmark, echo = FALSE}
br_sizes <- c(16, 32, 64, 128)

# ---- pipeflow: build once, time only forced re-runs ----
pips_pf <- lapply(br_sizes, make_branch_pip) |>
  stats::setNames(paste0("pf_br=", br_sizes))

pf_times <- lapply(pips_pf, function(p) {
  replicate(nrep, elapsed_time(pip_run(p, lgr = NULL, force = TRUE)))
})

# ---- targets: fresh tar_dir each iteration, only tar_make is timed ----
tar_times <- lapply(br_sizes, function(n) {
  replicate(nrep, {
    tar_dir({
      tar_script(make_branch_tar(n), ask = FALSE)
      elapsed_time(tar_make(callr_function = NULL, reporter = "silent"))
    })
  })
})

# Reshape into data.frames for plotting
pf_df <- do.call(rbind, lapply(seq_along(pf_times), function(i) {
  data.frame(expr = names(pf_times)[i], time = pf_times[[i]])
}))
tar_df <- do.call(rbind, lapply(seq_along(tar_times), function(i) {
  data.frame(expr = paste0("tar br=", br_sizes[i]), time = tar_times[[i]])
}))

d_pf  <- cbind(pf_df,  package = "pipeflow")
d_tar <- cbind(tar_df, package = "targets")
```

```{r dag-plot, echo = FALSE, fig.width = 7, fig.height = 4}
d_all <- rbind(d_pf, d_tar)
d_all$n <- as.numeric(gsub("[^0-9]", "", d_all$expr))

title <- sprintf("Fan-out DAG — %d iterations", nrep)
ggplot(d_all, aes(x = factor(n), y = time * 1000, fill = package)) +
  geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
  labs(y = "Time (ms)", x = "Number of branches", title = title) +
  theme_bw()
```


```{r, include = FALSE}
options(old)
```
