---
title: "pipeflow vs targets benchmark"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  md_document:
    variant: gfm
    preserve_yaml: true
params:
    nrep: 30
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

nrep <- params$nrep
old <- options(width = 100L)
```

```{r libs, include = FALSE}
devtools::load_all()
library(microbenchmark)
library(targets)
suppressPackageStartupMessages({
    library(readr)
    library(dplyr)
    library(ggplot2)
})
```


Package: pipeflow `r packageVersion("pipeflow")`, targets `r packageVersion("targets")`

## 1. Walkthrough example

Based on https://books.ropensci.org/targets/walkthrough.html

```{r helpers}
# General helpers
create_data_csv <- function(data = airquality, file = "data.csv") {
    utils::write.csv(data, file)
}
elapsed_time <- function(expr) {
    system.time(expr, gcFirst = FALSE)[["elapsed"]]
}


# Helper functions used in both pipelines
get_data <- function(file) {
    utils::read.csv(file) |> stats::na.omit()
}

fit_model <- function(data) {
    lm(Ozone ~ Temp, data) |> coefficients()
}

plot_model <- function(model, data) {
    ggplot(data) +
        geom_point(aes(x = Temp, y = Ozone)) +
        geom_abline(intercept = model[1], slope = model[2])
}
```


### pipeflow pipeline

```{r pipeflow-walkthrough-run}
tar_dir({
    # Setup data and pipeline
    create_data_csv(file = "data.csv")
    p <- pip_new("walkthrough") |>
        pip_add("data",  \(file = "data.csv") get_data(file)) |>
        pip_add("model", \(data = ~data) fit_model(data)) |>
        pip_add("plot",  \(model = ~model, data = ~data) plot_model(model, data))


    message("\nProof of principle full run (no skips)")
    pip_run(p)
    p_r <- replicate(nrep, elapsed_time(pip_run(p, lgr = NULL, force = TRUE)))

    message("\nProof of principle skipped run")
    pip_run(p)
    p_s <- replicate(nrep, elapsed_time(pip_run(p, lgr = NULL)))
})
```

### targets pipeline

```{r targets-walkthrough-run}
tar_make_here <- function(reporter = "silent") {
    tar_make(callr_function = NULL, reporter = reporter)
}

tar_dir({
    # Setup data and pipeline
    create_data_csv(file = "data.csv")
    tar_script({
        list(
            tar_target(file, "data.csv", format = "file"),
            tar_target(data, get_data(file)),
            tar_target(model, fit_model(data)),
            tar_target(plot, plot_model(model, data))
        )
    }, ask = FALSE)

    message("\nProof of principle full run (no skips)")
    tar_make_here(reporter = "timestamp")
    tar_r <- replicate(nrep, {
        tar_destroy(ask = FALSE)
        elapsed_time(tar_make_here(reporter = "silent"))
    })

    message("\nProof of principle skipped run")
    tar_make_here(reporter = "timestamp")
    tar_s <- replicate(nrep, {
        elapsed_time(tar_make_here(reporter = "silent"))
    })
})
```

### Runtimes

```{r walkthrough-runtime-table, echo = FALSE}
times <- list(p_r, p_s, tar_r, tar_s)
title <- sprintf("Walkthrough (4 steps) — %d iterations", nrep)

df <- data.frame(
    "Package"     = c("pipeflow","pipeflow","targets","targets"),
    "Operation"   = c("full run","skipped run","full run","skipped run"),
    "Median (ms)" = sapply(times, median) * 1000,
    "Mean (ms)"   = sapply(times, mean) * 1000,
    check.names = FALSE
)

knitr::kable(df, digits = 3, caption = title)
```

```{r walkthrough-runtime-plot, echo = FALSE, fig.width = 7, fig.height = 4}
data <- data.frame(
    "Package" = c("pipeflow", "targets") |> rep(each = 2 * nrep),
    "Operation" = c("full run", "skipped run") |> rep(each = nrep),
    "Time" = c(p_r, p_s, tar_r, tar_s) * 1000
)

ggplot(data, aes(x = Operation, y = Time, fill = Package)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    labs(y = "Time (ms)", x = NULL, title = title) +
    theme_bw()
```

## 2. Long linear pipeline (N sequential steps)

Each step depends on the output of the previous step: `s0 -> s1 -> s2 -> ... -> sN`.

```{r linear-helpers}
# Shared helpers for the linear pipeline — used by both pipeflow and targets
linear_s0   <- function() 0
linear_succ <- function(x) x + 1
```

```{r linear-pipeflow}
create_linear_pip <- function(n) {
  pip <- pip_new("linear") |> pip_add("s0", linear_s0)
  for (i in seq_len(n))
    pip_add(pip, paste0("s", i), \(x = ~ -1) linear_succ(x))
  pip
}

# Verify
p <- create_linear_pip(5)
pip_run(p)
stopifnot(p[["s10", "out"]] == 5)
```

```{r linear-targets}
create_linear_tar <- function(n) {
  targets <- c("  tar_target(s0, linear_s0())")
  for (i in seq_len(n))
    targets <- c(targets, sprintf("  tar_target(s%s, linear_succ(s%s))", i, i - 1))
  writeLines(c("library(targets)", "list(", paste(targets, collapse = ",\n"), ")"), "_targets.R")
}

# Verify
tar_dir({
  create_linear_tar(5)
  tar_make_here(reporter = "silent")
  stopifnot(tar_read(s5) == 5)
})
```

### Benchmark

```{r linear-benchmark, results='asis', echo = FALSE}
sizes <- c(16, 32, 64, 128)

# Pipeflow pipelines (built once, run repeatedly with force)
pips_pf <- lapply(sizes, create_linear_pip) |>
  stats::setNames(paste0("pf_n=", sizes))

# Run pipeflow benchmarks
mb_pf <- microbenchmark(
  control = list(order = "inorder"), times = nrep,
  "pf n=16"  = pip_run(pips_pf[["pf_n=16"]],  force = TRUE),
  "pf n=32"  = pip_run(pips_pf[["pf_n=32"]],  force = TRUE),
  "pf n=64"  = pip_run(pips_pf[["pf_n=64"]],  force = TRUE),
  "pf n=128" = pip_run(pips_pf[["pf_n=128"]], force = TRUE)
)

# Run targets benchmarks — fresh tar_dir per iteration
mb_tar <- microbenchmark(
  control = list(order = "inorder"), times = nrep,
  "tar n=16"  = tar_dir({ create_linear_tar(16);  tar_make_here() }),
  "tar n=32"  = tar_dir({ create_linear_tar(32);  tar_make_here() }),
  "tar n=64"  = tar_dir({ create_linear_tar(64);  tar_make_here() }),
  "tar n=128" = tar_dir({ create_linear_tar(128); tar_make_here() })
)

tab_pf  <- summary(mb_pf,  unit = "ms")[, c("expr", "median", "mean")]
tab_tar <- summary(mb_tar, unit = "ms")[, c("expr", "median", "mean")]

tab_pf[["us/step"]]  <- round(tab_pf[["median"]]  / sizes * 1000, 1)
tab_tar[["us/step"]] <- round(tab_tar[["median"]]  / sizes * 1000, 1)

tab <- rbind(tab_pf, tab_tar)
knitr::kable(tab, digits = 1)
```

```{r linear-plot, echo = FALSE, fig.width = 7, fig.height = 4}
d_pf  <- cbind(as.data.frame(mb_pf),  package = "pipeflow")
d_tar <- cbind(as.data.frame(mb_tar), package = "targets")

d_all <- rbind(d_pf, d_tar)
d_all$n <- as.numeric(gsub("[^0-9]", "", d_all$expr))

title <- sprintf("Linear pipeline — %d iterations", nrep)
ggplot(d_all, aes(x = factor(n), y = time / 1e6, fill = package)) +
  geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
  labs(y = "Time (ms)", x = "Number of steps", title = title) +
  theme_bw()
```

## 3. DAG with branching (fan-out: N parallel branches → 1 sink)

Source feeds N parallel branches, all converging on a single sink step.

```{r dag-helpers}
# Shared helpers for the fan-out DAG — used by both pipeflow and targets
dag_source <- function() 1
dag_branch <- function(x) x + 1
dag_sink   <- function(...) sum(...)
```

```{r dag-pipeflow}
make_branch_pip <- function(n) {
  pip <- pip_new("dag") |> pip_add("source", dag_source)
  for (i in seq_len(n))
    pip_add(pip, paste0("b", i), \(x = ~source) dag_branch(x))

  sink_args <- paste(sprintf("x%s = ~b%s", seq_len(n), seq_len(n)), collapse = ", ")
  sink_call <- paste(paste0("x", seq_len(n)), collapse = ", ")
  eval(parse(text = sprintf(
    "pip_add(pip, 'sink', function(%s) { dag_sink(%s) })", sink_args, sink_call
  )))
  invisible(pip)
}

p4 <- make_branch_pip(4); pip_run(p4)
stopifnot(p4[["sink", "out"]] == 8)
```

```{r dag-targets}
make_branch_tar <- function(n) {
  targets <- c("  tar_target(source, dag_source())")
  for (i in seq_len(n))
    targets <- c(targets, sprintf("  tar_target(b%s, dag_branch(source))", i))
  sink_vars <- paste(paste0("b", seq_len(n)), collapse = ", ")
  targets <- c(targets, sprintf("  tar_target(sink, dag_sink(%s))", sink_vars))
  writeLines(c("library(targets)", "list(", paste(targets, collapse = ",\n"), ")"), "_targets.R")
}

# Verify
tar_dir({
  make_branch_tar(4)
  tar_make_here()
  stopifnot(tar_read(sink) == 8)
})
```

### Benchmark

```{r dag-benchmark, results='asis', echo = FALSE}
br_sizes <- c(16, 32, 64, 128)

pips_pf <- lapply(br_sizes, make_branch_pip) |>
  stats::setNames(paste0("pf_br=", br_sizes))

mb_pf <- microbenchmark(
  control = list(order = "inorder"), times = 10,
  "pf br=16"  = pip_run(pips_pf[["pf_br=16"]],  force = TRUE),
  "pf br=32"  = pip_run(pips_pf[["pf_br=32"]],  force = TRUE),
  "pf br=64"  = pip_run(pips_pf[["pf_br=64"]],  force = TRUE),
  "pf br=128" = pip_run(pips_pf[["pf_br=128"]], force = TRUE)
)

mb_tar <- microbenchmark(
  control = list(order = "inorder"), times = 10,
  "tar br=16"  = tar_dir({ make_branch_tar(16);  tar_make_here() }),
  "tar br=32"  = tar_dir({ make_branch_tar(32);  tar_make_here() }),
  "tar br=64"  = tar_dir({ make_branch_tar(64);  tar_make_here() }),
  "tar br=128" = tar_dir({ make_branch_tar(128); tar_make_here() })
)

tab_pf  <- summary(mb_pf,  unit = "ms")[, c("expr", "median", "mean")]
tab_tar <- summary(mb_tar, unit = "ms")[, c("expr", "median", "mean")]

tab_pf[["us/node"]]  <- round(tab_pf[["median"]]  / (br_sizes + 2) * 1000, 1)
tab_tar[["us/node"]] <- round(tab_tar[["median"]]  / (br_sizes + 2) * 1000, 1)

tab <- rbind(tab_pf, tab_tar)
knitr::kable(tab, digits = 1)
```

```{r dag-plot, echo = FALSE, fig.width = 7, fig.height = 4}
d_pf  <- cbind(as.data.frame(mb_pf),  package = "pipeflow")
d_tar <- cbind(as.data.frame(mb_tar), package = "targets")

d_all <- rbind(d_pf, d_tar)
d_all$n <- as.numeric(gsub("[^0-9]", "", d_all$expr))

title <- sprintf("Fan-out DAG — %d iterations", nrep)
ggplot(d_all, aes(x = factor(n), y = time / 1e6, fill = package)) +
  geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
  labs(y = "Time (ms)", x = "Number of branches", title = title) +
  theme_bw()
```


## Session info

```{r session-info}
sessionInfo()
```

```{r, include = FALSE}
options(old)
```
