---
title: "pipeflow vs targets"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  A scoped latency comparison between pipeflow and targets.
vignette: >
  %\VignetteIndexEntry{pipeflow vs targets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  min_iters: 20
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#",
    prompt = FALSE,
    tidy = FALSE,
    cache = FALSE,
    collapse = TRUE
)

old <- options(width = 100L)
```

```{r libs, include = FALSE}
library(pipeflow)
library(targets)
library(bench)
library(ggplot2)

mi <- params$min_iters
```

## Overview

[{targets}](https://docs.ropensci.org/targets/) is the most widely used
pipeline toolkit in the R ecosystem and the de-facto standard for heavy-duty
reproducible workflows. The table below contrasts the two packages to help
you decide which one fits your project.

| Feature | **targets** | **pipeflow** |
|---|---|---|
| Paradigm | Declarative — define the full DAG upfront in a `_targets.R` script, then execute | Interactive — incrementally build the pipeline with `pip_add()` as you code |
| Execution | `tar_make()` runs in a **fresh R process** by default | `pip_run()` runs in the **current R session** |
| Persistent storage | ✅ Output stored to disk (`_targets/objects/`), survives R restarts, handles data larger than RAM | ❌ In-memory only |
| Skip up-to-date steps | ✅ Hash-based invalidation of code and data | ✅ State-based (`done` / `outdated`) |
| Metadata & provenance | ✅ `tar_meta()` records runtime, size, errors per target | ❌ No per-step provenance metadata (only the time of last execution) |
| Dependency validation | ✅ `tar_validate()` for pre-flight checks (opt-in) | ✅ On `pip_add()`, `pip_replace()`, `pip_remove()` — fails fast on broken references |
| Modify pipeline at runtime | ❌ Must edit `_targets.R` and re-run | ✅ `pip_add()` insert with `after =`, `pip_replace()`, `pip_remove()`  |
| Parameter management | ❌ No unified parameter view across targets | ✅ `pip_get_params()` / `pip_set_params()` — one call updates all steps |
| Split / map / reduce | ✅ `pattern = map()` / `cross()` built-in, `tarchetypes` for advanced patterns | ✅ Built-in `exec = "split"` / `"auto"` / `"reduce"` |
| Dynamic branching | ✅ Comprehensive via `tarchetypes` | ✅ Auto-mapping over partition keys (`exec = "auto"`) |
| Views / tag filtering | ✅ `tar_described_as()` selects by description tags | ✅ `pip_view()` and the `[` operator — filter by tags, step names, state, or boolean expressions |
| Pipeline composition | ❌ | ✅ `rbind()` any number of pipelines |
| Nested pipelines | ❌ | ✅ Steps can embed and run nested pipelines — reuse standalone pipelines as building blocks |
| Self-modifying pipelines | ❌ | ✅ Steps can modify the pipeline at runtime via `.self` (add/remove/replace steps, set params) and restart the run with `.self$restart()` |
| Distributed computing | ✅ `crew` for HPC and cloud workers | ❌ |
| Cloud storage | ✅ AWS, GCS | ❌ |
| File tracking | ✅ File targets with `format = "file"` | ❌ |
| Step locking | ❌ | ✅ `pip_lock()` / `pip_unlock()` — protect steps from accidental modification |
| Visualization / monitoring | ✅ `tar_visnetwork()`, `tar_glimpse()`, `tar_watch()` (live Shiny app) | ✅ `pip_graph()` → visNetwork |

In short, **{targets}** is the tool of choice for large-scale reproducible
projects: it persists results to disk, captures provenance with
`tar_meta()`, and scales to distributed infrastructure via `crew`.
**{pipeflow}** prioritises simplicity and interactivity: sub-millisecond
skipped-step checks, in-session pipeline modification, and low response
times make it well suited as a Shiny backend or for rapid parameter
exploration during analysis.

## The benchmark: what we measure

The comparison below is deliberately scoped. It does *not* claim that
pipeflow is generally "faster at pipelines" but rather tackle one
specific question:
*what is the in-session interactive latency of each package?*.
For this, we investigate how long does a running R session take to:

 - declare a pipeline
 - check whether anything needs to run, and
 - refresh the affected steps after a change

This covers the typical Shiny / parameter-exploration use case pipeflow
is built for.
For reference we also report the cold full end-to-end run, which is
not to be interpreted head-to-head but rather as a demonstration of the
different cost models of the two packages.

Each operation is timed as its own, clearly scoped phase:

| Phase | pipeflow | targets |
|---|---|---|
| **Define + build** | `pip_new()` + `pip_add()` loop | write `_targets.R`, then graph build (`tar_manifest()`) |
| **Skip-all** | `pip_run()` — all steps `done` | `tar_make()` — all targets up to date |
| **Warm incremental** | `pip_set_params()` + `pip_run()` | edit parameter in `_targets.R` + `tar_make(callr_function = NULL)` |
| **Cold full run** (reference) | `pip_run(force = TRUE)` | `tar_destroy()` + default `tar_make()` (fresh process) |

All phases run in the current R session, except the cold reference, where
targets uses its default fresh-process execution. targets re-sources
`_targets.R` and rebuilds its graph on *every* `tar_make()` and
`tar_manifest()` call — that fixed per-run cost is part of its in-session
latency and is measured here. pipeflow has no equivalent: its graph is
the in-memory object built once by `pip_add()`.

### Measurement method

* `bench::mark()` is used with `check = FALSE` (the two expressions return
  different objects by design), `memory = FALSE`, and `filter_gc = FALSE`.
* Each expression is evaluated once and discarded before timing, as an
  explicit warmup.
* since `bench::mark()` does *not* randomize expressions, every comparison
  is run twice, with the argument order swapped, and the per-iteration times
  pooled.
* Each expression runs for at least `mi` (= `r mi`) iterations and `min_time`
  seconds, capped at `max_iterations`.
* targets is run in the current session (`callr_function = NULL`) for all
  phases except the cold reference.
* The targets store lives in a fresh temporary directory (`tar_dir()`) on the
  same fast local disk for every run.
* Package versions and the full session are recorded at the end: pipeflow
  `r packageVersion("pipeflow")`, targets `r packageVersion("targets")`.

## The operations, in miniature

Before the timing, here are the pieces being benchmarked on a three-step
linear pipeline, to make the phases concrete.

```{r helpers}
# ---- pipeflow: linear pipeline --------------------------------------------
create_linear_pip <- function(n, flavor = "trivial", work = 0.005) {
    pip <- pip_new("linear") |> pip_add("s0", \(init = 0) init)
    for (i in seq_len(n)) {
        if (flavor == "costly") {
            pip_add(
                pip,
                step = sprintf("s%d", i),
                \(x = ~ -1) {
                    Sys.sleep(work)
                    x + 1
                }
            )
        } else {
            pip_add(pip, step = sprintf("s%d", i), \(x = ~ -1) x + 1)
        }
    }
    pip
}

# ---- targets: linear pipeline script --------------------------------------
linear_script <- function(n, flavor = "trivial", work = 0.005, init = 0) {
    lines <- c(
        "library(targets)",
        "list(",
        sprintf("    tar_target(s0, %s),", init)
    )
    for (i in seq_len(n)) {
        prev <- sprintf("s%d", i - 1)
        if (flavor == "costly") {
            body <- sprintf(
                paste0(
                    "    tar_target_raw(\"s%d\", quote({ Sys.sleep(%s);",
                    " %s + 1 }))"
                ),
                i, format(work, scientific = FALSE), prev
            )
        } else {
            body <- sprintf(
                "    tar_target_raw(\"s%d\", quote(%s + 1))", i, prev
            )
        }
        comma <- if (i == n) "" else ","
        lines <- c(lines, paste0(body, comma))
    }
    c(lines, ")")
}

# ---- shared ---------------------------------------------------------------
tar_make_silent <- function() {
    tar_make(callr_function = NULL, reporter = "silent")
}

# Each call flips between 0 and 1, so consecutive evaluations of the same
# package always see a genuinely new parameter value. One toggle per package,
# because the benchmark interleaves evaluations of the two packages.
counter <- function() {
    k <- 0L
    function() {
        k <<- 1L - k
        k
    }
}

# One discarded evaluation of each expression before timing.
warmup <- function(exprs, env) {
    for (e in exprs) {
        eval(e, env)
    }
    invisible(NULL)
}

# Turn a bench::mark() result into long (package, time[s]) data.
mark_times <- function(mark_res) {
    times <- mark_res$time
    data.frame(
        package = rep(as.character(mark_res$expression), times = lengths(times)),
        time = as.numeric(unlist(times)),
        stringsAsFactors = FALSE
    )
}

# Benchmark two named quoted expressions in both orders and pool the
# per-iteration times.
bench_pair <- function(exprs, env, min_iterations, min_time = 0.2,
                       max_iterations = 1000L) {
    stopifnot(length(exprs) == 2L, !is.null(names(exprs)))
    run <- function(e) {
        mark_times(bench::mark(
            exprs = e,
            env = env,
            check = FALSE,
            memory = FALSE,
            filter_gc = FALSE,
            min_iterations = min_iterations,
            min_time = min_time,
            max_iterations = max_iterations
        ))
    }
    rbind(run(exprs), run(exprs[rev(seq_along(exprs))]))
}

run_phase <- function(exprs, env, n, phase, min_iterations, min_time = 0.2,
                      max_iterations = 1000L) {
    warmup(exprs, env)
    d <- bench_pair(
        exprs, env,
        min_iterations, min_time = min_time, max_iterations = max_iterations
    )
    d$n <- n
    d$phase <- phase
    d
}

median_kable <- function(d, digits = 1) {
    agg <- aggregate(time * 1000 ~ package + n + phase, d, median)
    names(agg) <- c("package", "n", "phase", "median (ms)")
    knitr::kable(agg, digits = digits)
}
```

```{r verify-linear}
# pipeflow, trivial bodies
p <- create_linear_pip(3)
pip_run(p, lgr = NULL)
stopifnot(p[["s3", "out"]] == 3)

# pipeflow, costly bodies
p_costly <- create_linear_pip(3, flavor = "costly", work = 0.001)
pip_run(p_costly, lgr = NULL)
stopifnot(p_costly[["s3", "out"]] == 3)

# targets, trivial and costly bodies
tar_dir({
    writeLines(linear_script(3), "_targets.R")
    tar_make_silent()
    stopifnot(tar_read(s3) == 3)

    writeLines(linear_script(3, flavor = "costly", work = 0.001),
               "_targets.R")
    tar_make_silent()
    stopifnot(tar_read(s3) == 3)
})
```

The mini-pipelines build and run correctly in both packages, with trivial
and costly step bodies alike. Now watch what happens after a change at the
head of the chain:

```{r ops-demo}
# pipeflow: change one parameter, refresh affected steps
pip_set_params(p, list(init = 1))
print(p)
pip_run(p, lgr = NULL)

# pipeflow: everything up to date -> skip-all
pip_run(p, lgr = NULL)

# targets: edit the parameter value in _targets.R, refresh affected steps
tar_dir({
    writeLines(linear_script(3, init = 1), "_targets.R")
    tar_make_silent()
})
```

Both packages re-run only the steps affected by the change (the whole chain,
in a linear pipeline), and both skip everything when the pipeline is up to
date. These are the two operations we time in phases 2 and 3.

## Phase 1 — Define and build

How long does it take to declare a pipeline in R? For pipeflow, that is the
`pip_add()` loop itself. For targets, it is writing the target script plus
the graph build that `tar_manifest()` performs — and note that targets pays
this graph build on *every* `tar_make()` call, not just at definition time.

```{r define-benchmark, echo = FALSE}
sizes <- c(16, 32, 64, 128)

define_df <- do.call(rbind, lapply(sizes, function(n) {
    tar_dir({
        exprs <- list(
            pipeflow = quote(create_linear_pip(n)),
            targets = quote({
                writeLines(linear_script(n), "_targets.R")
                tar_manifest()
            })
        )
        run_phase(exprs, environment(), n, "define", 10L)
    })
}))
```

```{r define-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    define_df,
    aes(x = factor(n), y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    scale_y_log10() +
    labs(
        y = "Time (ms, log scale)",
        x = "Number of steps",
        title = "Phase 1: define + build"
    ) +
    theme_bw()
```

```{r define-medians, echo = FALSE}
median_kable(define_df)

define_med <- aggregate(time ~ package + n, define_df, median)
define_wide <- reshape(
    define_med,
    idvar = "n", timevar = "package", direction = "wide"
)
define_ratio <- round(range(define_wide$time.targets / define_wide$time.pipeflow))
```

Declaring the pipeline is roughly `r define_ratio[1]`–`r define_ratio[2]`
times faster in pipeflow across the four sizes, and the gap widens with `n`.
The targets bar covers writing `_targets.R` and the parse and graph build
that `tar_manifest()` performs — the entry price of its declarative model,
paid once here and again on every run.

## Phase 2 — Skip-all

The closest thing to an apples-to-apples comparison: everything is already
up to date, and we time a no-op run. This is the "sub-millisecond skip
check" claim, measured directly.

```{r skip-benchmark, echo = FALSE}
skip_df <- do.call(rbind, lapply(sizes, function(n) {
    tar_dir({
        writeLines(linear_script(n), "_targets.R")
        tar_make_silent()
        p <- create_linear_pip(n)
        pip_run(p, lgr = NULL)

        exprs <- list(
            pipeflow = quote(pip_run(p, lgr = NULL)),
            targets = quote(tar_make_silent())
        )
        run_phase(exprs, environment(), n, "skip", mi)
    })
}))
```

```{r skip-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    skip_df,
    aes(x = factor(n), y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    scale_y_log10() +
    labs(
        y = "Time (ms, log scale)",
        x = "Number of steps",
        title = "Phase 2: skip-all (nothing to do)"
    ) +
    theme_bw()
```

```{r skip-medians, echo = FALSE}
median_kable(skip_df)

skip_med <- aggregate(time ~ package + n, skip_df, median)
skip_pf_max_ms <- max(skip_med$time[skip_med$package == "pipeflow"]) * 1000
```

pipeflow's skip check stays at or below `r round(skip_pf_max_ms, 2)` ms even
at 128 steps. targets pays its per-run script re-source and graph rebuild
even when no target runs. That fixed cost is what a targets user pays on
*every* `tar_make()` in an interactive session — it buys the persistent,
provenance-tracking store.

## Phase 3 — Warm incremental update

The fair head-to-head for interactive work: a pipeline that is up to date, a
single parameter change, and the time until all affected steps are
refreshed. Each iteration changes the source value (pipeflow:
`pip_set_params()`; targets: editing the value in `_targets.R`), so both
packages re-run the same affected steps.

### Trivial step bodies

```{r incr-benchmark, echo = FALSE}
incr_df <- do.call(rbind, lapply(sizes, function(n) {
    tar_dir({
        scripts <- list(linear_script(n, init = 0L), linear_script(n, init = 1L))
        writeLines(scripts[[1L]], "_targets.R")
        tar_make_silent()
        p <- create_linear_pip(n)
        pip_run(p, lgr = NULL)

        pf_flip <- counter()
        tar_flip <- counter()
        exprs <- list(
            pipeflow = quote({
                pip_set_params(p, list(init = pf_flip()))
                pip_run(p, lgr = NULL)
            }),
            targets = quote({
                writeLines(scripts[[tar_flip() + 1L]], "_targets.R")
                tar_make_silent()
            })
        )
        run_phase(exprs, environment(), n, "incremental", mi)
    })
}))
```

```{r incr-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    incr_df,
    aes(x = factor(n), y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    labs(
        y = "Time (ms)",
        x = "Number of steps",
        title = "Phase 3: warm incremental update (trivial step bodies)"
    ) +
    theme_bw()
```

```{r incr-medians, echo = FALSE}
median_kable(incr_df)
```

With trivial step bodies this is pure framework overhead: the graph
traversal and state bookkeeping in pipeflow, plus the per-run graph rebuild
and per-step bookkeeping in targets. pipeflow's cost grows with the number
of re-run steps, while targets pays a fixed graph rebuild on top of a
heavier per-step cost.

### Work-cost control

Trivial bodies are overhead, but overhead is only meaningful relative to
real work. So we repeat the incremental scenario with a fixed cost per
step — `Sys.sleep(0.005)` — so each full cascade performs `n * 5 ms` of
known work. This isolates the framework's contribution to a run that
actually computes something.

```{r work-benchmark, echo = FALSE}
work_per_step <- 0.005

work_df <- do.call(rbind, lapply(sizes, function(n) {
    tar_dir({
        scripts <- list(
            linear_script(n, flavor = "costly", work = work_per_step, init = 0L),
            linear_script(n, flavor = "costly", work = work_per_step, init = 1L)
        )
        writeLines(scripts[[1L]], "_targets.R")
        tar_make_silent()
        p <- create_linear_pip(n, flavor = "costly", work = work_per_step)
        pip_run(p, lgr = NULL)

        pf_flip <- counter()
        tar_flip <- counter()
        exprs <- list(
            pipeflow = quote({
                pip_set_params(p, list(init = pf_flip()))
                pip_run(p, lgr = NULL)
            }),
            targets = quote({
                writeLines(scripts[[tar_flip() + 1L]], "_targets.R")
                tar_make_silent()
            })
        )
        run_phase(exprs, environment(), n, "incremental", 8L, min_time = 0.5)
    })
}))
```

```{r work-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    work_df,
    aes(x = factor(n), y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    labs(
        y = "Time (ms)",
        x = "Number of steps",
        title = paste0(
            "Phase 3: warm incremental update with ",
            work_per_step * 1000, " ms of work per step"
        )
    ) +
    theme_bw()
```

```{r work-medians, echo = FALSE}
work_med <- aggregate(time ~ package + n + phase, work_df, median)
work_med$median_ms <- work_med$time * 1000
work_med$work_ms <- work_med$n * work_per_step * 1000
work_med$overhead_ms <- work_med$median_ms - work_med$work_ms
work_med$share <- work_med$overhead_ms / work_med$median_ms
knitr::kable(
    work_med[c("package", "n", "median_ms", "work_ms", "overhead_ms", "share")],
    col.names = c("package", "n", "median (ms)", "work (ms)",
                  "overhead (ms)", "overhead share"),
    digits = c(0, 0, 1, 0, 1, 3)
)

work_wide <- reshape(
    work_med[, c("package", "n", "median_ms")],
    idvar = "n", timevar = "package", direction = "wide"
)
work_ratio <- max(work_wide$median_ms.targets / work_wide$median_ms.pipeflow)
```

Once steps do real work, the picture changes: the update time is
increasingly dominated by the work itself, and the two packages land within
about `r round(work_ratio, 1)`× of each other. The overhead share column makes the
same point quantitatively — it is a minority of the total for both. The
framework contribution is exactly what the trivial-bodies benchmark
isolates, which confirms that the benchmark measures interactive *latency*,
not pipeline throughput.

## Phase 4 — Cold full run (reference, not a race)

For completeness, here is the cold end-to-end rebuild. It is *not* a
head-to-head: pipeflow re-runs in a warm in-memory session, while targets
destroys its store and runs in a fresh R process, persisting every object
to disk with provenance. That is targets' design and its home turf — the
bar shows the magnitude of its cost model, not a defect.

This time the teardown is kept out of the timed region: `tar_destroy()` runs
before each timed `tar_make()`.

```{r cold-benchmark, echo = FALSE}
cold_n <- 128L
cold_iter <- 5L

p_cold <- create_linear_pip(cold_n)
pip_run(p_cold, lgr = NULL)
pf_cold <- vapply(seq_len(cold_iter), function(i) {
    gc(FALSE)
    system.time(
        pip_run(p_cold, lgr = NULL, force = TRUE),
        gcFirst = FALSE
    )[["elapsed"]]
}, numeric(1))

tar_cold <- tar_dir({
    writeLines(linear_script(cold_n), "_targets.R")
    vapply(seq_len(cold_iter), function(i) {
        tar_destroy(ask = FALSE)
        gc(FALSE)
        system.time(tar_make(reporter = "silent"), gcFirst = FALSE)[["elapsed"]]
    }, numeric(1))
})

cold_df <- data.frame(
    package = rep(c("pipeflow", "targets"), each = cold_iter),
    time = c(pf_cold, tar_cold),
    n = cold_n,
    phase = "cold"
)
```

```{r cold-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    cold_df,
    aes(x = package, y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3
    ) +
    scale_y_log10() +
    labs(
        y = "Time (ms, log scale)",
        x = NULL,
        title = "Phase 4: cold full run at 128 steps (reference only)"
    ) +
    theme_bw()
```

```{r cold-medians, echo = FALSE}
median_kable(cold_df)
```

## Does the story change with wide graphs?

A source feeds `n` parallel branches that converge on one sink. We time the
same two phases — skip-all and warm incremental — with trivial bodies.

```{r dag-helpers}
dag_source <- function() 1
dag_branch <- function(x) x + 1
dag_sink <- function(...) sum(...)

make_branch_pip <- function(n) {
    pip <- pip_new("dag") |>
        pip_add("source", \(init = 0) dag_source() + init)
    for (i in seq_len(n)) {
        pip_add(pip, sprintf("b%d", i),
                \(x = ~source) dag_branch(x))
    }
    args <- as.pairlist(stats::setNames(
        lapply(paste0("b", seq_len(n)), \(s) {
            stats::as.formula(paste0("~", s))
        }),
        paste0("x", seq_len(n))
    ))
    body <- as.call(c(
        list(as.symbol("dag_sink")),
        lapply(paste0("x", seq_len(n)), as.symbol)
    ))
    sink_fun <- eval(call("function", args, body))
    pip_add(pip, "sink", sink_fun)
}

branch_script <- function(n, init = 0) {
    branches <- vapply(seq_len(n), function(i) {
        sprintf(
            "    tar_target_raw(\"b%d\", quote(dag_branch(source))),",
            i
        )
    }, character(1))
    sink_line <- sprintf(
        "    tar_target_raw(\"sink\", quote(dag_sink(%s)))",
        paste(paste0("b", seq_len(n)), collapse = ", ")
    )
    c(
        "library(targets)",
        "dag_source <- function() 1",
        "dag_branch <- function(x) x + 1",
        "dag_sink <- function(...) sum(...)",
        "list(",
        sprintf("    tar_target(source, dag_source() + %s),", init),
        branches,
        sink_line,
        ")"
    )
}
```

```{r verify-dag}
p4 <- make_branch_pip(4)
pip_run(p4, lgr = NULL)
stopifnot(p4[["sink", "out"]] == 8)

tar_dir({
    writeLines(branch_script(4), "_targets.R")
    tar_make_silent()
    stopifnot(tar_read(sink) == 8)
})
```

```{r dag-benchmark, echo = FALSE}
dag_df <- do.call(rbind, lapply(sizes, function(n) {
    tar_dir({
        writeLines(branch_script(n), "_targets.R")
        tar_make_silent()
        p <- make_branch_pip(n)
        pip_run(p, lgr = NULL)

        skip <- run_phase(
            list(
                pipeflow = quote(pip_run(p, lgr = NULL)),
                targets = quote(tar_make_silent())
            ),
            environment(), n, "skip", mi
        )

        scripts <- list(branch_script(n, init = 0L), branch_script(n, init = 1L))
        pf_flip <- counter()
        tar_flip <- counter()
        incr <- run_phase(
            list(
                pipeflow = quote({
                    pip_set_params(p, list(init = pf_flip()))
                    pip_run(p, lgr = NULL)
                }),
                targets = quote({
                    writeLines(scripts[[tar_flip() + 1L]], "_targets.R")
                    tar_make_silent()
                })
            ),
            environment(), n, "incremental", mi
        )

        rbind(skip, incr)
    })
}))
```

```{r dag-plot, echo = FALSE, fig.width = 7, fig.height = 4}
ggplot(
    dag_df,
    aes(x = factor(n), y = time * 1000, fill = package)
) +
    geom_violin(trim = FALSE, alpha = 0.6, scale = "width") +
    stat_summary(
        fun = median,
        geom = "crossbar",
        width = 0.3,
        position = position_dodge(0.9)
    ) +
    scale_y_log10() +
    facet_wrap(~ phase, ncol = 2) +
    labs(
        y = "Time (ms, log scale)",
        x = "Number of branches",
        title = "Fan-out DAG: skip-all and warm incremental update"
    ) +
    theme_bw()
```

```{r dag-medians, echo = FALSE}
median_kable(dag_df)
```

Wide graphs do not change the picture: skip checks stay sub-millisecond in
pipeflow, and the incremental update scales with the number of re-run
steps while targets carries its fixed per-run rebuild.

## What the numbers include

The bars above are *not* all measuring the same thing. This table states
explicitly what each phase does and does not include.

| Bar | Step bodies run | Process | Disk persistence | Provenance | Store |
|---|---|---|---|---|---|
| pipeflow skip-all | none | current session | no | no | in-memory |
| targets skip-all | none | current session | yes | yes | on-disk `_targets/` |
| pipeflow incremental | affected steps | current session | no | no | in-memory |
| targets incremental | affected steps | current session | yes | yes | on-disk `_targets/` |
| pipeflow cold (ref) | all steps | current session | no | no | in-memory |
| targets cold (ref) | all steps | **fresh R process** | yes | yes | on-disk `_targets/` |

In short: targets' overhead buys disk persistence, provenance, and
distributed execution; pipeflow's buys in-session latency. The benchmark
above is a latency comparison, and the numbers only make sense with that
caveat attached.

## Summary

* **Skip-all** (Phase 2) is the closest apples-to-apples comparison and the
  strongest headline: pipeflow's skip check is sub-millisecond, targets'
  is dominated by its per-run graph rebuild.
* **Warm incremental** (Phase 3) is the fairest measure of interactive
  latency: pipeflow scales with the re-run steps, targets carries a fixed
  per-run cost, and with real per-step work the two converge.
* **Cold full run** (Phase 4) is reported as reference only: targets'
  fresh-process, disk-persisting rebuild is its home turf, not a race
  pipeflow can run.
* Treat these as *latency* numbers, not a general speed ranking. A targets
  run that saves hours of recomputation, or that runs across `crew`
  workers, will win the only race that matters for heavy workloads — and
  pipeflow will never claim that ground.

## Session info

```{r session-info, echo = FALSE}
sessionInfo()
```

```{r restore, include = FALSE}
options(old)
```
